---
title: "Lab 3: Logistic Regression"
author: "Chris Slaughter"
name: labs/lab3-logistic-questions.qmd
toc: true
---

## Introduction

In this lab, we will use logistic regression to analyze the salary dataset. We will consider the covariates gender and year of degree.

### Setup

Load packages

```{r}
#| echo: false
#| results: hide
library(rms)
library(ggplot2)
library(contingencytables) # For likelihood ratio test of 2x2 tables
library(lmtest) # For likelihood ratio tests with logistic regression
```

-   The contigencytables packages contains a function for calculating the likelihood ratio test for 2x2 tables

```{r}
perondi_2004
LR_test_2x2(perondi_2004)
```

## Part 1: Initial dataset manipulation

### 1. Read in the salary dataset

### 2. Remove all observations that are not from 1995

### 3. Create an indicator variable for male gender

### 4. Create a new outcome variable (salhigh) to indicate if salary is above \$7600 per month.

#### 4.1 What proportion of subjects have a salary above \$7600

## Part 2: High salary and gender (unadjusted model)

In part 2, we will estimate the association between salary and gender. First, we will conduct this analysis using the familiar method, a 2x2 table. We will then conduct the same analysis using logistic regression and compare the two approaches

### 1. Create a 2x2 contingency table of salhigh by male. From this table, calculate

#### 1.1. Pearson's Chi-squared test statistic (without continuity correction)

#### 1.2. The likelihood ratio Chi-squared test statistic

#### 1.3. The odds of high salary for males compared to the odds of a high salary for females (the odds ratio). Interpret

#### 1.4. The standard error of the log odds ratio (hint: the variance of the log odds ratio is given by 1/a + 1/b + 1/c + 1/d where a, b, c, and d are the cell totals)

#### 1.5. Pr(high salary \| male)

#### 1.6. Pr(high salary \| female)

### 2. Conduct a logistic regression of male on salhigh. From the regression output, calculate or identify

#### 2.1. The Wald Chi-squared test statistic (hint: Recall if Z is Standard Normal, the Z\^2 follow a Chi-squared distribution)

#### 2.2. The likelihood ratio Chi-squared statistic

#### 2.3. The odds of high salary for males compared to the odds of a high salary for females (the odds ratio). Interpret the odds ratio.

#### 2.4. The standard error of the log odds ratio

#### 2.5. Pr(high salary \| male)

#### 2.6. Pr(high salary \| female)

### 3. Obtain estimates for the predicted probabilities 2.5 and 2.6 (use Stata or R commands to get these numbers)

## Part 3: High salary and year of degree (unadjusted model)

### 1. Create suitable summary statistics for the probability of high salary by year of degree

### 2. Fit a logistic regression model with salhigh and yrdeg

#### 2.1. Calculate the odds ratio and interpret the results

### 3. Suppose the investigators would rather model the number of years it has been since each professor has achieved his or her degree

#### 3.1. Create a new variable yrssince95 = 95 - yrdeg

#### 3.2. Fit a logistic regression model with salhigh and yrssince95

#### 3.3. Compare the output from model 3.2 with 2.1

### 4. From model 3.2, create a variable for the predicted probability of having a high salary.

#### 4.1. Calculate the odds and log odds of having a high salary

#### 4.2. Plot yrssince95 versus the log odds of a having a high salary

#### 4.3. Plot yrssince95 versus the odds of having a high salary

#### 4.4. Plot yrssince95 versus the probability of having a high salary

### 5. Conduct a likelihood ratio test to test the statistical significance of yrssince95

#### 5.1. Fit a logistic regression model with just the intercept. Save the deviance from this model

#### 5.2. Fit logistic regression model 3.2. Save the deviance from this model.

#### 5.3. Subtract the deviance obtained from 5.2 from the deviance obtained from 5.1. Compare this result to the likelihood ratio test statistic given in the output from model 3.2 (and 2.1)
