---
title: "Lab 8 Supplement: Regression Models for Contingency Tables"
subtitle: "Binary, Ordinal, Multinomial, and Poisson Approaches"
format: 
   html:
     toc: true
     toc-depth: 3
---

## Overview

This lab examines a single $3 \times 3$ contingency table through the lens of four
different regression models: binary logistic, proportional odds ordinal logistic,
multinomial logistic, and Poisson log-linear. Each model makes different assumptions
about the outcome, yet all draw inference from the same underlying table of counts.
The goals are to:

1. Determine which models are **saturated** (i.e., have enough parameters to
   reproduce every cell probability exactly).
2. **Calculate intercepts and slopes by hand** from the cell counts and verify
   they match the fitted model output.
3. **Discuss the advantages and disadvantages** of each modeling approach.

We will also consider the role of covariate adjustment (age and CHD) and close
with a Poisson log-linear formulation that unifies contingency table analysis
with rate modeling and survival analysis.

## Data Setup

The data form a $3 \times 3$ table with a 3-level outcome `y` (0, 1, 2) and a
3-level predictor `x` (0, 1, 2).

```{r}
#| label: data-setup

# Recreate the dataset from the cell counts
d1 <- expand.grid(y = 0:2, x = 0:2, wt = NA)
d1$wt <- c(30, 15, 45, 18, 12, 60, 10, 8, 72)

# Expanded (one-row-per-subject) dataset
d2 <- data.frame(
  x = rep(d1$x, d1$wt),
  y = rep(d1$y, d1$wt)
)

# Verify the table (reversed so highest levels appear top-left)
table(factor(d2$x, levels = 2:0), factor(d2$y, levels = 2:0))
```

We also create dichotomized and factor versions of the variables that will be
needed across models.

```{r}
#| label: derived-vars

d2$y.bin   <- (d2$y >= 1) + 0
d2$x.bin   <- (d2$x >= 1) + 0
d2$x.factor <- factor(d2$x)
d2$y.factor <- factor(d2$y)
```

A utility function for computing logits by hand:

```{r}
#| label: mylogit

mylogit <- function(p) log(p / (1 - p))
```


## Model 1: Binary Logistic Regression

Here we collapse the 3-level outcome to binary ($Y \ge 1$ vs.\ $Y = 0$) and
the predictor to binary ($X \ge 1$ vs.\ $X = 0$). This yields a $2 \times 2$
table.

```{r}
#| label: binary-model

library(rms)

# Fit via lrm
lrm(y.bin ~ x.bin, data = d2)

# Fit via glm (for comparison)
glm(y.bin ~ x.bin, family = "binomial", data = d2)
```

### Hand Calculations

```{r}
#| label: binary-hand

# Marginal counts for the collapsed 2x2 table
table(factor(d2$y.bin, levels = 1:0), factor(d2$x.bin, levels = 1:0))

# Intercept = logit(Pr(Y>=1 | X=0))
# Among X=0: 60 events out of 90
mylogit(60 / 90)

# Slope = log odds ratio
# OR = (a*d) / (b*c) from the 2x2 table
# X=0: Y>=1 has 60, Y=0 has 30
# X=1: Y>=1 has 152, Y=0 has 28  (pooling x=1 and x=2)
log(30 * 152 / (28 * 60))
```

### Saturation

The binary logistic model with a single binary predictor has **2 parameters**
(intercept + slope) for a $2 \times 2$ table that has **2 free probabilities**
(one per row, since probabilities sum to 1 within rows). So it is saturated
for the collapsed $2 \times 2$ table — it reproduces those cell proportions
exactly.

But the question of interest is whether the model is saturated **for the
original $3 \times 3$ table**, and the answer is **no**. The $3 \times 3$
table has $3 \times 2 = 6$ free conditional probabilities (2 per row). The
binary logistic model estimates only 2 parameters. By collapsing both the
outcome and predictor from 3 levels to 2, we have discarded information about
the internal structure of the table. The model cannot recover the individual
cell probabilities of the $3 \times 3$ table — it only describes a coarsened
summary. Whether that loss matters depends on whether the collapsed categories
are scientifically meaningful.


## Model 2: Multinomial Logistic Regression

The multinomial model treats all 3 outcome levels as nominal categories, using
$Y = 0$ as the reference. With `x.factor` (3 levels), the model has a separate
set of log-odds for each non-reference outcome level at each predictor level.

**Note:** The multinomial model uses a **log link** on the ratio of category
probabilities (i.e., log relative risk or log-odds relative to the reference
category), not a logit link. Each equation models
$\log[\Pr(Y = j \mid X) / \Pr(Y = 0 \mid X)]$ as a linear function of $X$.

* For $k=1$,

$$
\log[\Pr(Y = 1 \mid X=x) / \Pr(Y = 0 \mid X=x)] = \beta_{01} + \beta_{11}I(X=1) + \beta_{21}I(X=2)
$$
* Isolate $\beta_{21}$

* $x=2$

$$
\log[\Pr(Y = 1 \mid X=2) / \Pr(Y = 0 \mid X=2)] = \beta_{01} + \beta_{21}
$$
* $x=0$

$$
\log[\Pr(Y = 1 \mid X=0) / \Pr(Y = 0 \mid X=0)] = \beta_{01}
$$
$$
\beta_{21} = \log[\Pr(Y = 1 \mid X=2) / \Pr(Y = 0 \mid X=2)] - \log[\Pr(Y = 1 \mid X=0) / \Pr(Y = 0 \mid X=0)] = \beta_{01}

$$


```{r}
#| label: multinomial-model

library(nnet)

m_multi <- multinom(y ~ x.factor, data = d2)
summary(m_multi)
```

### Hand Calculations

The multinomial intercepts and slopes can be computed directly from the cell
counts. For outcome level $j$ relative to $Y = 0$:

$$
\beta_{j0} = \log\left(\frac{n_{0j}}{n_{00}}\right), \qquad
\beta_{jk} = \log\left(\frac{n_{kj}}{n_{k0}}\right) - \log\left(\frac{n_{0j}}{n_{00}}\right)
$$

where $n_{kj}$ is the count in predictor level $k$, outcome level $j$.

```{r}
#| label: multinomial-hand

# Slopes for Y=2 (relative to Y=0)
# beta_21 (x=1 effect): log(n_12/n_10) - log(n_02/n_00)
log(72/90 / (30/90)) - log(45/90 / (30/90))

# Slopes for Y=1 (relative to Y=0)
log(8/90 / (10/90)) - log(15/90 / (30/90))

# Equivalently, beta_21 is a log odds ratio from the 2x2 sub-table
# comparing (x=2, y=1) vs (x=0, y=0):
with(d2[d2$x %in% c("0", "2") & d2$y %in% c("0", "1"), ],
     table(factor(x, levels = c("2","0")), factor(y, levels = c("1","0"))))
log(8 * 30 / (15 * 10))  # log OR


# This would be the easier way to obtain the other betas, e.g. beta_22
with(d2[d2$x %in% c("0", "2") & d2$y %in% c("0", "2"), ],
     table(factor(x, levels = c("2","0")), factor(y, levels = c("2","0"))))
log(72 * 30 / (45 * 10))  # log OR
```

### Saturation

With 3 predictor levels and 3 outcome levels, the table has $3 \times 2 = 6$
free probabilities (2 per row, since the 3 outcome probabilities sum to 1
within each row). The multinomial model with `x.factor` estimates
$2 \times 3 = 6$ parameters (an intercept and two slopes for each of the 2
non-reference outcome levels). The model is **saturated** — it will always
reproduce the observed cell proportions exactly, regardless of the data.


## Model 3: Proportional Odds Ordinal Logistic Regression

The proportional odds model treats the outcome as ordinal and assumes a
**common odds ratio** across all binary cutpoints of $Y$.

```{r}
#| label: ordinal-model

m_ord <- lrm(y ~ x.factor, data = d2, x = TRUE)
m_ord
```

### Hand Calculations

The proportional odds model has two intercepts (one for each cutpoint of $Y$)
and, with `x.factor`, two slopes ($X = 1$ and $X = 2$ effects). But the key
constraint is that each slope is the **same** across cutpoints.

We can verify the proportional odds assumption by checking whether the log-odds
ratios are the same at each cutpoint:

```{r}
#| label: ordinal-hand

# At cutpoint Y >= 1 vs Y = 0:
mylogit(30/90) - mylogit(10/90)

# At cutpoint Y >= 2 vs Y <= 1:
mylogit(45/90) - mylogit(18/90)

# These should be equal if proportional odds holds
# (and in this dataset, they are — by construction)
```

### Saturation

The proportional odds model with `x.factor` (3 levels) has $2 + 2 = 4$
parameters (2 intercepts + 2 slopes). But the $3 \times 3$ table has 6 free
probabilities. So this model is **not saturated** in general — it has 2
residual degrees of freedom.

However, in this particular dataset, proportional odds holds exactly (the
log-odds ratios are identical across cutpoints), so the model happens to
reproduce the cell probabilities perfectly. This is a property of **this
dataset**, not of the model itself.

```{r}
#| label: ordinal-vs-multinomial-predictions

# Predicted probabilities from both models — they agree here
newdata <- data.frame(x.factor = levels(d2$x.factor))

cat("Multinomial predicted probabilities:\n")
predict(m_multi, newdata = newdata, "probs")

cat("\nOrdinal predicted probabilities:\n")
predict(m_ord, newdata, type = "fitted")
```

The fact that these predictions agree is because $2\hat\beta_1 = \hat\beta_2$
happens to hold in the ordinal model.  This is a coincidence of this dataset, not a
general result.


## Adjusting for Covariates

### Age

*Should we adjust for age?*

Key considerations: Is age a confounder? Is it
associated with both $X$ and $Y$? If so, adjustment may be warranted to
obtain a less biased estimate of the $X$–$Y$ association. However, if the
study is designed so that $X$ is randomized or balanced on age, adjustment
may not be necessary (age could be a precision variable, which in the case of logistic regression tends to drive associations away from the null)

### CHD (Coronary Heart Disease)

*Should we adjust for CHD?*

Key considerations: Is CHD a mediator on the
causal pathway from $X$ to $Y$? If so, adjusting for it would block part
of the causal effect and bias the total effect estimate. Is CHD a collider?
Adjusting for a collider can introduce spurious associations. The decision
depends on the causal structure.  A DAG is helpful here.


## Model 4: Poisson Log-Linear Regression

### Motivation

So far, we have modeled the **conditional distribution** of $Y$ given $X$
using logistic, ordinal, and multinomial models. Each conditions on the row
margins (the number of subjects at each level of $X$) and models the
probability of falling into each outcome category.

There is an alternative: instead of conditioning on the margins, we can model
the **cell counts directly** using Poisson regression. This is the log-linear
model approach to contingency tables.

Why would we do this? Several reasons:

1. It provides a **unified framework** that encompasses all the previous models
   as special cases or constrained versions.
2. It connects contingency table analysis to **rate modeling** and
   **survival analysis** through the concept of offsets.
3. It makes the saturated model explicit — the interaction terms in the
   Poisson model are exactly the association parameters.


### The Log-Linear Model

We reshape the $3 \times 3$ table into a "long" dataset with one row per cell
(9 rows), a count variable, and factor variables for $X$ and $Y$.

The saturated Poisson log-linear model is:

$$
\log(\mu_{ij}) = \lambda + \alpha_i + \beta_j + \gamma_{ij}
$$

where:

- $\lambda$ is an overall intercept (related to total sample size)
- $\alpha_i$ captures the **row main effects** (predictor distribution)
- $\beta_j$ captures the **column main effects** (outcome distribution)
- $\gamma_{ij}$ captures the **interaction** (the association between $X$ and $Y$)

The interaction terms $\gamma_{ij}$ are the parameters of scientific interest.
They encode the same information as the log-odds ratios from the multinomial
model.

```{r}
#| label: poisson-setup

# Cell-level data for the Poisson model
d_cells <- d1
d_cells$x.factor <- factor(d_cells$x)
d_cells$y.factor <- factor(d_cells$y)
d_cells$count    <- d_cells$wt

d_cells
```

```{r}
#| label: poisson-saturated

# Saturated Poisson log-linear model
m_pois <- glm(count ~ x.factor * y.factor, family = poisson, data = d_cells)
summary(m_pois)
```

### Interpreting the Parameters

The main effects for `x.factor` capture how the row totals differ — they
absorb the marginal distribution of the predictor. The main effects for
`y.factor` capture how the column totals differ — they absorb the marginal
distribution of the outcome. Neither set of main effects tells us about the
**association** between $X$ and $Y$.

The interaction terms are where the action is. Each interaction parameter
is a **log odds ratio** comparing a specific ($X$, $Y$) cell to the
reference cell, relative to what would be expected under independence.

```{r}
#| label: poisson-vs-multinomial

# Compare interaction parameters to multinomial slopes
cat("Poisson interaction coefficients:\n")
coef(m_pois)[grep(":", names(coef(m_pois)))]

cat("\nMultinomial coefficients:\n")
coef(m_multi)
```

The Poisson interaction terms and the multinomial regression coefficients
encode the **same log-odds ratios** — they are just organized differently.
The multinomial model parameterizes them as slopes within each outcome
equation; the Poisson model parameterizes them as interaction terms in a
single equation.


### Why Does This Work? The Poisson–Multinomial Equivalence

This equivalence is sometimes called the **Birch result** or the
**Poisson trick**. The key insight:

If you model cell counts as independent Poisson random variables with a
log-linear model, and then condition on the observed row totals (the marginal
counts of $X$), the conditional distribution of the cell counts within each
row is **multinomial**. Since the row main effects $\alpha_i$ perfectly absorb
the row margins, the Poisson likelihood factors into a part that depends only
on the main effects and a part that depends only on the interaction — and the
interaction part is exactly the multinomial likelihood.

In practice, this means:

- The interaction parameter estimates are **identical** whether you fit a
  Poisson log-linear model or a multinomial logistic model.
- The standard errors and test statistics for the interaction terms are also
  identical.
- The row and column main effects are "bonus" parameters that the multinomial
  model doesn't estimate (because it conditions on the margins).


### Saturation of the Poisson Model

The saturated Poisson log-linear model for a $3 \times 3$ table has
$3 \times 3 = 9$ parameters (intercept + 2 row effects + 2 column effects +
4 interactions) for 9 cell counts. It is **saturated** — it reproduces every
cell count exactly.

```{r}
#| label: poisson-fitted

# Fitted values should exactly match observed counts
cbind(observed = d_cells$count, fitted = fitted(m_pois))
```

If we drop the interaction terms (the independence model), we have
$1 + 2 + 2 = 5$ parameters for 9 cells, leaving 4 residual degrees of
freedom. This reduced model tests the hypothesis of **independence** between
$X$ and $Y$ — it is the log-linear analogue of the chi-squared test.

```{r}
#| label: poisson-independence

# Independence model (no interaction)
m_pois_indep <- glm(count ~ x.factor + y.factor, family = poisson, data = d_cells)
summary(m_pois_indep)

# Likelihood ratio test for independence
anova(m_pois_indep, m_pois, test = "LRT")
```


### The Role of Offsets (and Why We Don't Need One Here)

In many applications of Poisson regression, an **offset** term appears:

$$
\log(\mu_i) = \log(t_i) + \mathbf{x}_i'\boldsymbol{\beta}
$$

where $t_i$ is the exposure (person-time, population size, etc.). The offset
standardizes the counts so that we model a **rate** ($\mu_i / t_i$) rather
than a raw count. Without the offset, we would confound "more events" with
"more opportunity to observe events."

**In our contingency table, no offset is needed.** Why?

Every subject contributes exactly one observation — they enter the study, get
classified into one cell, and that is it. There is no variation in the
opportunity to be counted. Every person had the same "exposure": one unit of
observation, one chance to land in a cell.

The implicit offset is $\log(1) = 0$ for every cell, and so it vanishes from
the model. The row and column main effects handle the marginal totals, which
is the contingency table analogue of what an offset does in the rate context.

To make this concrete, consider what happens if we *do* use an offset equal
to the log of the row totals:

```{r}
#| label: poisson-with-offset

# With offset = log(row total), we model conditional probabilities
d_cells$row_total <- ave(d_cells$count, d_cells$x.factor, FUN = sum)

m_pois_offset <- glm(count ~ x.factor * y.factor + offset(log(row_total)),
                      family = poisson, data = d_cells)

# The interaction terms are unchanged
cat("Without offset (interaction terms):\n")
coef(m_pois)[grep(":", names(coef(m_pois)))]

cat("\nWith offset (interaction terms):\n")
coef(m_pois_offset)[grep(":", names(coef(m_pois_offset)))]
```

The interaction terms — the **association parameters** — are identical with or
without the offset. The offset only changes the main effect estimates, which
are nuisance parameters for the association question. When the offset equals
the log row total, the model explicitly conditions on the row margins and the
main effects now represent the conditional log-probabilities rather than the
marginal log-counts. But the association is the same either way.

This is the Birch result in action: inference about association is invariant
to whether you treat the margins as fixed (offset / multinomial) or random
(Poisson with main effects).


<!-- ### Connection to Survival Analysis -->

<!-- The Poisson log-linear model for contingency tables is not just an academic -->
<!-- curiosity — it is the **same model** that underlies rate-based survival -->
<!-- analysis. The only thing that changes is what goes in the offset. -->

<!-- Consider the unified model: -->

<!-- $$ -->
<!-- \log(\mu_i) = \log(t_i) + \mathbf{x}_i'\boldsymbol{\beta} -->
<!-- $$ -->

<!-- The question is: what is $t_i$? -->

<!-- | Setting                          | What is $t_i$?                                                    | Offset                     | -->
<!-- |----------------------------------|-------------------------------------------------------------------|----------------------------| -->
<!-- | **Contingency table**            | 1 for everyone (one classification per subject)                   | $\log(1) = 0$ (vanishes)  | -->
<!-- | **Rate table with person-time**  | Person-time in cell $i$, varies across cells                      | $\log(t_i)$, explicit     | -->
<!-- | **Piecewise exponential model**  | Person-time contributed by subject $i$ within a time interval     | $\log(t_i)$, explicit     | -->

<!-- In the piecewise exponential survival model, each subject's follow-up is -->
<!-- split into discrete time intervals, and within each interval the hazard is -->
<!-- assumed constant. The data are restructured into a table where each row -->
<!-- represents a subject–interval combination, the outcome is the event indicator -->
<!-- (0 or 1), and the offset is the log of the person-time contributed to that -->
<!-- interval. This is a Poisson regression — the same log-linear model we are -->
<!-- using here. -->

<!-- **The contingency table is the simplest special case.** Every subject has the -->
<!-- same exposure ($t_i = 1$), so the offset vanishes. The moment exposure varies -->
<!-- — because people are censored at different times, or because you are -->
<!-- aggregating person-time into intervals — you need the offset, and you are -->
<!-- in the piecewise exponential framework. -->

<!-- The progression is: -->

<!-- 1. **Contingency table** (equal exposure, no offset) — what we have here -->
<!-- 2. **Rate table** (unequal exposure, offset needed) — e.g., incidence rate -->
<!--    data where person-time varies across cells -->
<!-- 3. **Piecewise exponential model** (individual-level varying exposure) — -->
<!--    each subject contributes different person-time to different intervals -->
<!-- 4. **Cox model** (limiting case) — as the time intervals become -->
<!--    infinitesimally fine, the piecewise exponential model converges to the -->
<!--    Cox proportional hazards model; the partial likelihood is the limit of -->
<!--    the Poisson full likelihood -->

<!-- This means the Cox model — which can seem like a fundamentally different kind -->
<!-- of object — is actually the end of a continuum that starts with the simple -->
<!-- Poisson log-linear model for a contingency table. The partial likelihood is -->
<!-- what you get when you let the baseline hazard (the analogue of the column -->
<!-- main effects $\beta_j$) become nonparametric. -->

<!-- For this lab, the key takeaway is: **the equal-exposure assumption is what -->
<!-- makes the contingency table simple.** Every subject gets one classification -->
<!-- event with the same amount of "time" to accumulate it. That assumption is -->
<!-- trivially true here, which is why we don't need an offset. But naming this -->
<!-- invisible assumption prepares you to recognize when it fails — which is -->
<!-- exactly when you need offsets, rate models, and survival analysis. -->


### Robust Standard Errors and the Mean–Variance Assumption

The Poisson distribution assumes that the variance equals the mean:
$\text{Var}(Y_i) = \mu_i$. In many applications of Poisson regression —
particularly with individual-level count data — this assumption is suspect.
Counts often exhibit **overdispersion** (variance exceeding the mean) due to
unobserved heterogeneity, clustering, or other sources of extra-Poisson
variation. In those settings, the model-based standard errors from `glm()`
are too small, and it is common practice to use **robust (sandwich) standard
errors** to obtain valid inference without relying on the variance assumption.

**In the contingency table setting, this concern does not arise.** Here is why:
the saturated Poisson log-linear model has one parameter per cell count —
9 parameters for 9 observations. The fitted values reproduce the observed
counts exactly. There are zero residual degrees of freedom and no residual
deviance. The mean–variance relationship is not testable because there is no
residual variation left to assess. The model fits the data perfectly by
construction.

```{r}
#| label: poisson-dispersion

# Residual deviance and df for the saturated model
cat("Residual deviance:", deviance(m_pois), "\n")
cat("Residual df:", df.residual(m_pois), "\n")

# For comparison, the independence model does have residual df
# and we can check for overdispersion there
cat("\nIndependence model:\n")
cat("Residual deviance:", deviance(m_pois_indep), "\n")
cat("Residual df:", df.residual(m_pois_indep), "\n")
cat("Dispersion estimate:", deviance(m_pois_indep) / df.residual(m_pois_indep), "\n")
```

For the **independence model** (which is not saturated), overdispersion could
in principle be relevant — if the residual deviance substantially exceeds the
residual degrees of freedom, the Poisson variance assumption may be violated.
In that case, one could either use `family = quasipoisson` (which estimates a
dispersion parameter) or compute robust standard errors using the sandwich
estimator.

```{r}
#| label: poisson-robust

# Robust SEs via sandwich estimator (for the independence model)
library(sandwich)
library(lmtest)

cat("Model-based SEs (independence model):\n")
coeftest(m_pois_indep)

cat("\nRobust (sandwich) SEs:\n")
coeftest(m_pois_indep, vcov = vcovHC(m_pois_indep, type = "HC0"))
```

The robust standard errors protect against misspecification of the variance
function. When the mean model is correct but the variance assumption is wrong,
the point estimates from Poisson regression remain consistent (they solve the
correct estimating equation), but the standard errors need correction. This is
the same logic as using robust SEs in ordinary least squares when
heteroscedasticity is present.

For the **saturated model**, robust standard errors are not needed — the model
is exact. But the concept becomes important when you move to non-saturated
Poisson models (e.g., with continuous predictors, or the independence model)
or to the rate and survival settings discussed above, where the Poisson
variance assumption is a real constraint. The sandwich estimator is the
standard tool for relaxing that constraint while keeping the convenient
Poisson regression framework.


## Summary: Comparing the Four Models

```{r}
#| label: summary-table
#| echo: false

summ <- data.frame(
  Model = c("Binary Logistic", "Proportional Odds", "Multinomial", "Poisson Log-Linear"),
  `Outcome Treatment` = c(
    "Dichotomized (Y >= 1 vs Y = 0)",
    "Ordinal (cumulative logits)",
    "Nominal (3 categories, Y=0 ref)",
    "Cell counts (9 cells)"
  ),
  `Predictor Treatment` = c(
    "Binary (X >= 1 vs X = 0)",
    "Factor (3 levels)",
    "Factor (3 levels)",
    "Factor (3 levels)"
  ),
  Parameters = c(2, 4, 6, 9),
  `Free Cells` = c(2, 6, 6, 9),
  Saturated = c("Yes (for 2x2 only; No for 3x3)", "No (in general)", "Yes", "Yes"),
  check.names = FALSE
)

knitr::kable(summ, caption = "Comparison of the four regression approaches")
```

### Advantages and Disadvantages

**Binary Logistic:**
Simple, familiar, and well-understood. But collapsing the outcome loses
information — we can no longer distinguish between outcome levels 1 and 2.
The choice of where to dichotomize is often arbitrary and can affect
conclusions.

**Proportional Odds:**
Preserves the ordinal structure and is parsimonious (fewer parameters). The
proportional odds assumption constrains the model, which is a strength when
the assumption holds (more power, simpler interpretation) and a weakness when
it does not (biased estimates, poor fit). Not saturated for $> 2$ predictor
levels, so model fit should be checked.

**Multinomial:**
Makes no assumptions about the outcome structure — fully flexible. Always
saturated when the predictor is categorical. But this flexibility comes at a
cost: more parameters to estimate, and no borrowing of strength across outcome
levels. With continuous predictors or sparse data, the multinomial model can
be unstable.

**Poisson Log-Linear:**
Provides a unified framework that connects to rate modeling and survival
analysis. The saturated model is equivalent to the multinomial. The
independence model provides a global test of association. The offset mechanism
generalizes naturally to settings with unequal exposure. The main
disadvantage for contingency table analysis per se is that the parameterization
(main effects + interactions) is less directly interpretable than the
conditional probabilities of the multinomial model — though the interaction
terms are the same log-odds ratios.
