---
title: "Bootstrap Applications in Regression"
name: Bootstrap.qmd
---

```{r setup, message=FALSE, warning=FALSE}
library(Hmisc)
library(ggplot2)
library(knitr)
library(kableExtra)
tryCatch(source('pander_registry.R'), error = function(e) invisible(e))
```

# Introduction

These notes accompany [Set14.pdf](Set14.pdf), which covers bootstrap methodology including the theoretical foundations, bias correction, and confidence interval construction. Here, we apply bootstrap methods to prediction problems from Lecture 13:

- Compute **bootstrap confidence intervals for marginal means** in the beta carotene trial
- Compare the precision of unadjusted group means vs. ANCOVA marginal means using bootstrap CIs
- Discuss the choice of bootstrap scheme when predictors are **fixed by design** (as in a clinical trial with pre-specified dose levels)

Students should read [Set14.pdf](Set14.pdf) first for the theoretical background.

# Fixed vs. Random Predictors and the Bootstrap

## Why the distinction matters

In the beta carotene trial, subjects were randomly assigned to one of five dose levels (0, 15, 30, 45, or 60 mg/day). The dose levels and the number of subjects per dose were determined by the study design — they are **fixed**, not randomly sampled from a population.

This has two important consequences:

1.  **Standard errors**: Classical (model-based) standard errors are appropriate when $X$ is fixed and the variance model is correctly specified. Robust (sandwich) standard errors protect against heteroscedasticity, but when $X$ is fixed by design and the model is correctly specified, they offer no improvement over classical SEs — they simply lose efficiency. In Lectures 12 and 13, we used classical SEs for this reason.

2.  **Bootstrap scheme**: The standard "pairs" bootstrap (resampling entire rows of data) implicitly treats $X$ as random — each bootstrap sample may have different numbers of subjects per dose group. For a designed experiment with fixed $X$, we should instead use a bootstrap that conditions on the design.

## Two bootstrap approaches for fixed X

**Stratified bootstrap**: Resample subjects *within* each dose group, preserving the group sample sizes exactly. This respects the fixed-$X$ design directly.

**Residual bootstrap**: Fit the model, extract residuals, resample the residuals, and construct new responses $y^* = \hat{y} + e^*$. This conditions on $X$ by keeping the predictor values fixed and only resampling the noise.

We demonstrate both below and compare the results.

# Bootstrap CIs for Marginal Means

## Setup: Beta Carotene Trial

In Lecture 13, we estimated marginal means (population-average predicted plasma beta-carotene at 9 months) for each dose group using the ANCOVA model with a threshold + linear parameterization. The marginal mean at dose $d$ is:

$$
\hat{\mu}_{\text{marg}}(d) = \frac{1}{n} \sum_{i=1}^{n} \hat{y}_i(d, x_{0i})
$$

where $x_{0i}$ is subject $i$'s observed baseline value. Because this is an average of predictions over the empirical distribution of baseline values, classical standard error formulas (which apply to linear functions of $\hat{\beta}$) do not directly apply. The bootstrap provides a straightforward solution.

```{r}
carot <- Hmisc::stata.get("data/carot.dta")
dose.levels <- sort(unique(carot$dose))

# Threshold + linear model from Lecture 12/13
carot$trt <- (carot$dose > 0) + 0
m.thres <- lm(carot3 ~ trt + dose + carot0, data = carot)

# Unadjusted model for comparison
m.unadj <- lm(carot3 ~ factor(dose), data = carot)
```

## Approach 1: Stratified Bootstrap

We resample subjects with replacement *within each dose group*, so that every bootstrap sample has the same number of subjects per dose as the original study.

```{r, cache=TRUE}
set.seed(42)
B <- 2000

# Indices for each dose group
dose.idx <- split(seq_len(nrow(carot)), carot$dose)

boot.marginal.strat <- matrix(NA, nrow = B, ncol = length(dose.levels))
boot.unadj.strat    <- matrix(NA, nrow = B, ncol = length(dose.levels))

for(b in 1:B) {
  # Resample within each dose group
  idx <- unlist(lapply(dose.idx, function(ii) sample(ii, replace = TRUE)))
  boot.dat <- carot[idx, ]

  # Marginal means from ANCOVA
  m.b <- lm(carot3 ~ trt + dose + carot0, data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd <- data.frame(dose = dose.levels[j],
                     carot0 = carot$carot0,
                     trt = (dose.levels[j] > 0) + 0)
    boot.marginal.strat[b, j] <- mean(predict(m.b, newdata = nd))
  }

  # Unadjusted group means
  m.u <- lm(carot3 ~ factor(dose), data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd.u <- data.frame(dose = dose.levels[j])
    boot.unadj.strat[b, j] <- predict(m.u, newdata = nd.u)
  }
}
```

## Approach 2: Residual Bootstrap

We keep the original predictor values fixed and resample only the residuals.

```{r, cache=TRUE}
set.seed(42)

# Residuals from the fitted models
resid.thres <- residuals(m.thres)
resid.unadj <- residuals(m.unadj)
fitted.thres <- fitted(m.thres)
fitted.unadj <- fitted(m.unadj)

boot.marginal.resid <- matrix(NA, nrow = B, ncol = length(dose.levels))
boot.unadj.resid    <- matrix(NA, nrow = B, ncol = length(dose.levels))

for(b in 1:B) {
  # Resample residuals and construct new responses
  e.star.thres <- sample(resid.thres, replace = TRUE)
  e.star.unadj <- sample(resid.unadj, replace = TRUE)

  boot.dat <- carot
  boot.dat$carot3.thres <- fitted.thres + e.star.thres
  boot.dat$carot3.unadj <- fitted.unadj + e.star.unadj

  # ANCOVA marginal means
  m.b <- lm(carot3.thres ~ trt + dose + carot0, data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd <- data.frame(dose = dose.levels[j],
                     carot0 = carot$carot0,
                     trt = (dose.levels[j] > 0) + 0)
    boot.marginal.resid[b, j] <- mean(predict(m.b, newdata = nd))
  }

  # Unadjusted group means
  m.u <- lm(carot3.unadj ~ factor(dose), data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd.u <- data.frame(dose = dose.levels[j])
    boot.unadj.resid[b, j] <- predict(m.u, newdata = nd.u)
  }
}
```

## Comparing the Two Approaches

```{r}
# Point estimates (same for both approaches)
marginal.pts <- sapply(dose.levels, function(d) {
  nd <- data.frame(dose = d, carot0 = carot$carot0, trt = (d > 0) + 0)
  mean(predict(m.thres, newdata = nd))
})
unadj.pts <- predict(m.unadj, newdata = data.frame(dose = dose.levels))

# CIs from both approaches
ci.marg.strat <- data.frame(
  Lower = apply(boot.marginal.strat, 2, quantile, 0.025),
  Upper = apply(boot.marginal.strat, 2, quantile, 0.975))
ci.marg.resid <- data.frame(
  Lower = apply(boot.marginal.resid, 2, quantile, 0.025),
  Upper = apply(boot.marginal.resid, 2, quantile, 0.975))

comparison <- data.frame(
  Dose = dose.levels,
  Estimate = round(marginal.pts, 1),
  `Stratified CI` = sprintf("(%.1f, %.1f)", ci.marg.strat$Lower, ci.marg.strat$Upper),
  `Residual CI` = sprintf("(%.1f, %.1f)", ci.marg.resid$Lower, ci.marg.resid$Upper),
  `Strat Width` = round(ci.marg.strat$Upper - ci.marg.strat$Lower, 1),
  `Resid Width` = round(ci.marg.resid$Upper - ci.marg.resid$Lower, 1),
  check.names = FALSE
)

kable(comparison,
      caption = "ANCOVA marginal means: stratified vs. residual bootstrap 95% CIs") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   Both approaches condition on $X$ (the dose design) and should give similar results

-   The residual bootstrap assumes the residuals are exchangeable (approximately equal variance across groups). If there is substantial heteroscedasticity, the stratified bootstrap is more robust

-   For the remainder of this analysis, we use the stratified bootstrap results

## Precision Comparison: Unadjusted vs. ANCOVA Marginal

In Lecture 13, we noted that the theoretical variance reduction from ANCOVA is proportional to $(1 - \rho^2)$. We can now verify this empirically by comparing bootstrap CI widths.

```{r}
# Unadjusted CIs (from stratified bootstrap)
unadj.ci <- data.frame(
  Dose = dose.levels,
  Estimate = as.numeric(unadj.pts),
  Lower = apply(boot.unadj.strat, 2, quantile, 0.025),
  Upper = apply(boot.unadj.strat, 2, quantile, 0.975)
)

# Marginal CIs (from stratified bootstrap)
marginal.ci <- data.frame(
  Dose = dose.levels,
  Estimate = marginal.pts,
  Lower = ci.marg.strat$Lower,
  Upper = ci.marg.strat$Upper
)

# Side-by-side comparison
precision <- data.frame(
  Dose = dose.levels,
  `Unadjusted Width` = unadj.ci$Upper - unadj.ci$Lower,
  `Marginal Width` = marginal.ci$Upper - marginal.ci$Lower,
  check.names = FALSE
)
precision$`Ratio (Marginal / Unadj)` <- precision$`Marginal Width` /
  precision$`Unadjusted Width`

kable(precision, digits = c(0, 1, 1, 3),
      caption = "Bootstrap 95% CI widths: unadjusted vs. ANCOVA marginal means") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
rho <- cor(carot$carot0, carot$carot3)
sprintf("Baseline-outcome correlation: rho = %.3f", rho)
sprintf("Theoretical CI width ratio: sqrt(1 - rho^2) = %.3f", sqrt(1 - rho^2))
```

-   The empirical CI width ratio should be close to $\sqrt{1 - \rho^2}$, the theoretical standard error reduction factor from ANCOVA

-   The marginal means and unadjusted means estimate the same population quantity (the average outcome at each dose), but the ANCOVA-based marginal means are more precise because they leverage the baseline-outcome correlation

```{r, fig.width=10, fig.height=6, fig.cap="Unadjusted group means (gray) vs. ANCOVA marginal means (green) with bootstrap 95% CIs. Both estimate the same quantity; the marginal means are more precise."}
plot(NULL, xlim = c(-3, 63), ylim = c(-200, 2800),
     xlab = "Dose (mg/day)", ylab = "Plasma Beta-Carotene at 9 Months",
     main = "Precision Gain from ANCOVA: Bootstrap CIs")

# Unadjusted
arrows(unadj.ci$Dose - 0.8, unadj.ci$Lower, unadj.ci$Dose - 0.8, unadj.ci$Upper,
       angle = 90, code = 3, length = 0.04, col = "gray50", lwd = 2)
points(unadj.ci$Dose - 0.8, unadj.ci$Estimate, pch = 19, col = "gray50", cex = 1.5)

# ANCOVA marginal
arrows(marginal.ci$Dose + 0.8, marginal.ci$Lower, marginal.ci$Dose + 0.8, marginal.ci$Upper,
       angle = 90, code = 3, length = 0.04, col = "darkgreen", lwd = 2)
points(marginal.ci$Dose + 0.8, marginal.ci$Estimate, pch = 15, col = "darkgreen", cex = 1.5)

legend("topleft",
  c("Unadjusted group means", "ANCOVA marginal means"),
  col = c("gray50", "darkgreen"), pch = c(19, 15), lwd = 2, bty = "n")
```
