---
title: "Bootstrap Applications in Regression"
subtitle: "Supplementary Notes"
name: Bootstrap.Supplement.qmd
---

```{r setup, message=FALSE, warning=FALSE}
library(rms)
library(Hmisc)
library(ggplot2)
library(knitr)
library(kableExtra)
tryCatch(source('pander_registry.R'), error = function(e) invisible(e))
```

# Introduction

These notes supplement the bootstrap lecture, which is the primary reference for bootstrap methodology including the theoretical foundations, bias correction, and confidence interval construction. The purpose of these supplementary notes is to:

- Revisit the beta carotene example from Lecture 13 and compute **bootstrap confidence intervals for marginal means**
- Compare the precision of unadjusted group means vs. ANCOVA marginal means using bootstrap CIs
- Carry out **bootstrap validation** of the logistic regression model for predicting smoking status (the `lrm()` case study from Lecture 13)
- Demonstrate the `validate()` and `calibrate()` functions from the `rms` package

Students should read the bootstrap lecture notes first and treat these notes as a companion that applies bootstrap methods to prediction problems introduced in Lecture 13.

# Bootstrap CIs for Marginal Means

## Setup: Beta Carotene Trial

In Lecture 13, we estimated marginal means (population-average predicted plasma beta-carotene at 9 months) for each dose group using the ANCOVA model with a threshold + linear parameterization. The marginal mean at dose $d$ is:

$$
\hat{\mu}_{\text{marg}}(d) = \frac{1}{n} \sum_{i=1}^{n} \hat{y}_i(d, x_{0i})
$$

where $x_{0i}$ is subject $i$'s observed baseline value. Because this is an average of predictions over the empirical distribution of baseline values, classical standard error formulas (which apply to linear functions of $\hat{\beta}$) do not directly apply. The bootstrap provides a straightforward solution.

```{r}
carot <- Hmisc::stata.get("data/carot.dta")
dose.levels <- sort(unique(carot$dose))

# Threshold + linear model from Lecture 12/13
carot$trt <- (carot$dose > 0) + 0
m.thres <- lm(carot3 ~ trt + dose + carot0, data = carot)
```

## Bootstrap Procedure

For each bootstrap replicate:

1.  Resample $n$ observations with replacement from the original data
2.  Refit the threshold + linear ANCOVA model
3.  For each dose level, compute the marginal mean by averaging predictions over the **original** baseline distribution (not the bootstrap baseline distribution, which would add unnecessary noise)

```{r, cache=TRUE}
set.seed(42)
B <- 2000

boot.marginal <- matrix(NA, nrow = B, ncol = length(dose.levels))
boot.unadj    <- matrix(NA, nrow = B, ncol = length(dose.levels))

for(b in 1:B) {
  idx <- sample(nrow(carot), replace = TRUE)
  boot.dat <- carot[idx, ]

  # Marginal means from ANCOVA
  m.b <- lm(carot3 ~ trt + dose + carot0, data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd <- data.frame(dose = dose.levels[j],
                     carot0 = carot$carot0,
                     trt = (dose.levels[j] > 0) + 0)
    boot.marginal[b, j] <- mean(predict(m.b, newdata = nd))
  }

  # Unadjusted group means
  m.u <- lm(carot3 ~ factor(dose), data = boot.dat)
  for(j in seq_along(dose.levels)) {
    nd.u <- data.frame(dose = dose.levels[j])
    boot.unadj[b, j] <- predict(m.u, newdata = nd.u)
  }
}
```

## Bootstrap CIs for Marginal Means

```{r}
# Point estimates
marginal.pts <- sapply(dose.levels, function(d) {
  nd <- data.frame(dose = d, carot0 = carot$carot0, trt = (d > 0) + 0)
  mean(predict(m.thres, newdata = nd))
})

# Percentile bootstrap CIs
marginal.ci <- data.frame(
  Dose = dose.levels,
  Estimate = marginal.pts,
  Lower = apply(boot.marginal, 2, quantile, 0.025),
  Upper = apply(boot.marginal, 2, quantile, 0.975)
)

kable(marginal.ci, digits = 1,
      col.names = c("Dose (mg/day)", "Marginal Mean", "2.5%", "97.5%"),
      caption = "ANCOVA marginal means with bootstrap 95% CIs") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   These are the confidence intervals that were deferred from Lecture 13, where only point estimates were shown

## Precision Comparison: Unadjusted vs. ANCOVA Marginal

In Lecture 13, we noted that the theoretical variance reduction from ANCOVA is proportional to $(1 - \rho^2)$. We can now verify this empirically by comparing bootstrap CI widths.

```{r}
# Unadjusted point estimates and CIs
m.unadj <- lm(carot3 ~ factor(dose), data = carot)
unadj.pts <- predict(m.unadj, newdata = data.frame(dose = dose.levels))

unadj.ci <- data.frame(
  Dose = dose.levels,
  Estimate = unadj.pts,
  Lower = apply(boot.unadj, 2, quantile, 0.025),
  Upper = apply(boot.unadj, 2, quantile, 0.975)
)

# Side-by-side comparison
comparison <- data.frame(
  Dose = dose.levels,
  `Unadjusted Width` = unadj.ci$Upper - unadj.ci$Lower,
  `Marginal Width` = marginal.ci$Upper - marginal.ci$Lower,
  check.names = FALSE
)
comparison$`Ratio (Marginal / Unadj)` <- comparison$`Marginal Width` /
  comparison$`Unadjusted Width`

kable(comparison, digits = c(0, 1, 1, 3),
      caption = "Bootstrap 95% CI widths: unadjusted vs. ANCOVA marginal means") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
rho <- cor(carot$carot0, carot$carot3)
sprintf("Baseline-outcome correlation: rho = %.3f", rho)
sprintf("Theoretical CI width ratio: sqrt(1 - rho^2) = %.3f", sqrt(1 - rho^2))
```

-   The empirical CI width ratio should be close to $\sqrt{1 - \rho^2}$, the theoretical standard error reduction factor from ANCOVA

-   The marginal means and unadjusted means estimate the same population quantity (the average outcome at each dose), but the ANCOVA-based marginal means are more precise because they leverage the baseline-outcome correlation

```{r, fig.width=10, fig.height=6, fig.cap="Unadjusted group means (gray) vs. ANCOVA marginal means (green) with bootstrap 95% CIs. Both estimate the same quantity; the marginal means are more precise."}
plot(NULL, xlim = c(-3, 63), ylim = c(-200, 2800),
     xlab = "Dose (mg/day)", ylab = "Plasma Beta-Carotene at 9 Months",
     main = "Precision Gain from ANCOVA: Bootstrap CIs")

# Unadjusted
arrows(unadj.ci$Dose - 0.8, unadj.ci$Lower, unadj.ci$Dose - 0.8, unadj.ci$Upper,
       angle = 90, code = 3, length = 0.04, col = "gray50", lwd = 2)
points(unadj.ci$Dose - 0.8, unadj.ci$Estimate, pch = 19, col = "gray50", cex = 1.5)

# ANCOVA marginal
arrows(marginal.ci$Dose + 0.8, marginal.ci$Lower, marginal.ci$Dose + 0.8, marginal.ci$Upper,
       angle = 90, code = 3, length = 0.04, col = "darkgreen", lwd = 2)
points(marginal.ci$Dose + 0.8, marginal.ci$Estimate, pch = 15, col = "darkgreen", cex = 1.5)

legend("topleft",
  c("Unadjusted group means", "ANCOVA marginal means"),
  col = c("gray50", "darkgreen"), pch = c(19, 15), lwd = 2, bty = "n")
```

# Bootstrap Validation of a Logistic Regression Model

## Setup: Smoking Prediction Model

In Lecture 13, we built a logistic regression model to predict smoking status from the FEV dataset using Harrell's 4-step framework. Steps 1--3 (data assessment, data reduction, model fitting) were carried out in Lecture 13. Here we complete **Step 4: Model evaluation** using bootstrap validation.

```{r}
fev <- read.csv(file = "data/FEV.csv")
fev$smoker <- as.integer(fev$smoke == "current smoker")

dd <- datadist(fev)
options(datadist = "dd")

# Refit the model from Lecture 13
f <- lrm(smoker ~ sex * rcs(age, 4) + rcs(height, 4) + fev,
         data = fev, x = TRUE, y = TRUE)
```

## Bootstrap Validation with `validate()`

The `validate()` function from `rms` performs bootstrap cross-validation:

1.  Draw a bootstrap sample (sample with replacement)
2.  Fit the model to the bootstrap sample
3.  Evaluate performance on the bootstrap sample ("training") and the original sample ("test")
4.  The difference is the **optimism** — the amount by which apparent performance overstates true performance
5.  Repeat $B$ times and average

```{r, cache=TRUE}
v <- validate(f, B = 300)
v
```

-   Key rows in the `validate()` output:

```{r}
val.guide <- data.frame(
  Index = c("Dxy", "R2", "Slope", "B"),
  Description = c(
    "Somers' Dxy rank correlation (C = 0.5 + Dxy/2)",
    "Nagelkerke R-squared",
    "Calibration slope (1.0 = perfectly calibrated)",
    "Brier score (lower is better)"
  ),
  `Original` = c("Apparent performance on full data",
                  "Apparent performance on full data",
                  "Always 1.0 on training data",
                  "Apparent Brier score"),
  `Corrected` = c("Honest estimate after removing optimism",
                   "Honest estimate after removing optimism",
                   "Degree of overfitting (< 1 means overfit)",
                   "Honest Brier score"),
  check.names = FALSE
)
kable(val.guide, caption = "Guide to key validate() output") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   The `optimism` column is the difference between `training` and `test` performance across bootstrap samples

    -   Large optimism indicates the model is overfitting the data

    -   The `index.corrected` = `index.orig` $-$ `optimism` gives the honest estimate

## Calibration with `calibrate()`

Calibration asks: when the model predicts a 30% probability of smoking, do approximately 30% of such patients actually smoke?

```{r, fig.cap="Bootstrap calibration plot for the smoking prediction model", cache=TRUE}
cal <- calibrate(f, B = 300)
plot(cal, main = "Bootstrap Calibration: Smoking Prediction Model")
```

-   The calibration plot shows predicted probability (x-axis) versus observed event rate (y-axis)

    -   A well-calibrated model follows the 45-degree line

    -   The gray region shows the confidence band from bootstrap resampling

    -   Departures from the line indicate systematic over- or under-prediction at certain risk levels

## Summary: Discrimination vs. Calibration

-   **Discrimination** tells you whether the model can *rank* patients correctly (measured by the C-statistic)

-   **Calibration** tells you whether the predicted *probabilities* are accurate (measured by the calibration slope and calibration plot)

-   A model can have good discrimination but poor calibration (or vice versa). Both matter for clinical use.

-   The bootstrap provides a principled way to estimate both quantities honestly, correcting for the optimism inherent in evaluating a model on its training data
