---
title: "Poisson Regression"
subtitle: "Lecture 09"
name: Lec09.Poisson.qmd
---

```{r}
#| echo: false
library(rms)
library(ggplot2)
library(splines)
tryCatch(source('pander_registry.R'), error = function(e) invisible(e))
```

## Count Data and Event Rates

-   Sometimes a random variable measures the number of events occurring over some space and time interval

-   Examples include

    -   Number of polyps recurring in the three year interval between colonoscopies

    -   Number of pulmonary exacerbations experienced by a cystic fibrosis patient in a year

    -   Number of reflux events in a 24-hour period

-   Count data have (in theory) no upper limit, although very large counts can be highly improbable

-   When a response variable measures counts over space and time, we often summarize by considering the event rate

    -   "Event rate" is the expected number of events per unit of space-time

    -   The rate is thus a mean count

    -   In most statistical problems, we know the interval of time and the volume of space sampled

        -   Poisson models allow us to take into account the known interval of time/space using an "offset"

## Poisson Model

### Poisson distribution

-   Often we assume that counts follow a Poisson distribution

-   The Poisson distribution can be derived from the following assumptions

    -   The expected number of events in an interval is proportional to the size of the interval

    -   The probability that two events occur with an infinitesimally small interval of space-time is zero

    -   The number of events occurring in disjoint (separate) intervals of space-time are independent

-   (Note that the assumption of a constant rate with independence over space-time is pretty strong and rarely holds completely)

-   Poisson distribution

    -   Counts the events occurring at a constant rate $\lambda$ in a specified time (and space) $t$

        -   Independent intervals of time and space

    -   Probability distribution has parameter $\lambda > 0$

        -   For $k = 0, 1, 2, \ldots$

            $$\textrm{Pr}(Y = k) = \frac{e^{-\lambda t} (\lambda t)^k}{k!}$$

        -   Mean: $E[Y] = \lambda t$

        -   Var: $V[Y] = \lambda t$

        -   (Mean-variance relationship, like binary data)

### Regression Model

-   When the response variable represent counts of some event, we usually model using the (log) rate with Poisson regression

    -   Compares rates of response per space-time (e.g. person-years) across groups

    -   "Rate ratio"

-   Why not use linear regression? The reasons are primarily statistical

    -   The rate is in fact a mean

    -   For Poisson $Y$ having event rate $\lambda$ measured over time $t$

        -   The mean is equal to the variance (both are $\lambda t$)

    -   We want to be able to account for

        -   Different areas of space or length of time for measuring counts

        -   Mean-variance relationship (if not using robust standard errors)

-   In Poisson regression, we tend to use a log link when modeling the event rate

    -   As in other models, a log link means that we are assuming a multiplicative modeling

        -   Multiplicative model $\rightarrow$ comparisons between groups based on ratios

        -   Additive model $\rightarrow$ comparisons between groups based on differences

    -   Log link also has the best technical statistical properties

        -   Log rate is the "canonical parameter" for the Poisson distribution

        -   Being the canonical parameter makes the calculus and mathematical properties easier to derive, and thus easier to understand from a theoretical perspective

-   Poisson regression

    -   Response variable is count of event over space-time (often person-years)

    -   Offset variable specifies amount of space-time

    -   Allows continuous or multiple grouping variables

        -   But will also work with binary grouping variables

-   Simple Poisson Regression

    -   Modeling rate of count response $Y$ on predictor $X$

        -   Distribution: $\textrm{Pr}(Y_i = k | T_i = t_i) = \frac{e^{-\lambda_i t_i} (\lambda_i t_i)^k}{k!}$

        -   Model: $\textrm{log } E[Y_i | T_i, X_i] = \textrm{log}\left(\lambda_i T_i\right) = \textrm{log}(T_i) + \beta_0 + \beta_1 \times X_i$

        -   $X_i = 0$: log $\lambda_i = \beta_0$

        -   $X_i = x$: log $\lambda_i = \beta_0 + \beta_1 \times x$

        -   $X_i = x+1$: log $\lambda_i = \beta_0 + \beta_1 \times x + \beta_1$

    -   To interpret as rates, exponentiate the parameters

        -   Distribution: $\textrm{Pr}(Y_i = k | T_i = t_i) = \frac{e^{-\lambda_i t_i} (\lambda_i t_i)^k}{k!}$

        -   Model: $\textrm{log } E[Y_i | T_i, X_i] = \textrm{log}\left(\lambda_i T_i\right) = \textrm{log}(T_i) + \beta_0 + \beta_1 \times X_i$

        -   $X_i = 0$: $\lambda_i = e^{\beta_0}$

        -   $X_i = x$: $\lambda_i = e^{\beta_0 + \beta_1 \times x}$

        -   $X_i = x+1$: $\lambda_i = e^{\beta_0 + \beta_1 \times x + \beta_1}$

-   Interpretation of the model

    -   Intercept

        -   Rate when the predictor is $0$ is found by exponentiation of the intercept from Poisson regression: $e^{\beta_0}$

    -   Slope

        -   Rate ratio between groups differing in the value of the predictor by 1 unit is found by exponentiation of the slope from Poisson regression: $e^{\beta_1}$

## Example: Acid reflux and BMI

### Data description

-   Research question: Are the number of acid reflux events in a day related to body mass index (BMI)?

-   Each subject pH in the esophagus in monitored continuously for about 24 hours

-   Count the number of time pH drop below 4, which is called a "reflux event"

-   Analysis (statistical) goals

    -   Primary goal: Determine if there is an association between BMI and acid reflux rate

    -   Secondary goal: Describe the (mean) trend in reflux rates as a function of BMI

-   Variables

    -   Response: Number of acid reflux events

    -   Offset: Number of minutes subject was monitored

    -   Predictor of interest: BMI

    -   Other covariates: Presence of esophagitis at baseline

### Descriptive Plots

```{r, fig.cap="BMI by number of reflux events"}
bmi.data <- read.csv("data/bmi.csv", header=TRUE)
# Events are pH less than 4
bmi.data$events <- bmi.data$totalmins4

ggplot(bmi.data, aes(x=bmi, y=events)) + geom_point() + geom_smooth() + theme_bw() + xlab("BMI") + ylab("Number of reflux events")
```

```{r, fig.cap="BMI by number of reflux events per day"}
bmi.data$rate <- bmi.data$totalmins4/bmi.data$totalmins*60*24
ggplot(bmi.data, aes(x=bmi, y=rate)) + geom_point() + geom_smooth() + theme_bw() + xlab("BMI") + ylab("Reflux events per day")

```

-   Characterization of plots

    -   Plots are visually similar if we consider the rate (events per day) or the raw number of events

    -   First order trend: Event rate increases with increasing BMI

    -   Second order trend: Event rate increase until BMI of 32 (or so) and then flattens out

    -   Within-group variability

        -   Hard to visualize from the plots

        -   Model assumes increasing variability with increasing BMI, which looks reasonable

### Regression commands

-   As before, but need to specify the offset

    -   Offset is the log of the exposure time

    -   In Stata, can alternatively specify the "exposure" and it will take the log for you

-   Stata

    -   `poisson respvar predvar, exposure(time) [robust]`

    -   `poisson respvar predvar, offset(logtime) [robust]`

-   R

    -   One method to fit Poisson models

        -   Uses the `sandwich` and `lmtest` libraries

        -   Must install the above two libraries using `install.packages("lmtest")` and `install.packages("sandwich")`

        -   `model.poisson <- glm(response ~ predictors + offset(log(time)), data=data, family="poisson")`

        -   `coeftest(model.poisson, vcov=sandwich)`

    -   Another method to fit Poisson models using the `Design` package

        -   `m1 <- glmD(response ~ predictors + offset(log(time)), data=data, family="poisson", x=TRUE, y=TRUE)`

        -   `bootcov(m1)` for robust (bootstrap) confidence intervals

    -   Can also use methods within the `gee` library

### Estimation of the regression model

-   Regression model for number of reflux events on BMI

    -   Answer primary research question: Is there an *association* between BMI and the acid reflux event rate?

    -   Estimate the best fitting line to (log) number of reflux events within BMI groups using an offset of log time

        -   $\textrm{log}(\textrm{Events} | \textrm{BMI}) = \beta_0 + \beta_1 \times \textrm{BMI} + \textrm{log}(\textrm{time})$

    -   An association will exist if the slope $\beta_1$ is nonzero

```{r}
library(lmtest)
library(sandwich)
m1.poisson <- glm(events ~ bmi + offset(log(totalmins)), data=bmi.data, family="poisson")
m1.poisson
coeftest(m1.poisson, vcov=sandwich)
confint(coeftest(m1.poisson, vcov=sandwich))
```

``` stata
. poisson events bmi, offset(logmins) robust

Iteration 0:   log pseudolikelihood =  -11360.89  
Iteration 1:   log pseudolikelihood =  -11360.89  

Poisson regression                                Number of obs   =        279
                                                  Wald chi2(1)    =      23.42
                                                  Prob > chi2     =     0.0000
Log pseudolikelihood =  -11360.89                 Pseudo R2       =     0.0520

------------------------------------------------------------------------------
             |               Robust
      events |      Coef.   Std. Err.      z    P>|z|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
         bmi |   .0223194   .0046121     4.84   0.000     .0132799    .0313589
       _cons |  -3.119991    .139521   -22.36   0.000    -3.393448   -2.846535
     logmins |   (offset)
------------------------------------------------------------------------------
```

-   Interpretation of output

    -   $\textrm{log rate} = -3.119991 + 0.0223194 \times \textrm{BMI}$

-   Interpretation of intercept

    -   Estimated event rate when BMI is 0 is found by exponentiation: $e^{-3.12} = 0.044$

    -   This is the rate per 2-minute interval. This unusual time interval is an artifact of the way in pH data is sampled

        -   To convert to events per day, multiply by 720 (there are 720 2-minute intervals in a day)

        -   $720 \times e^{-3.12} = 31.7 \textrm{ events per day}$

-   Interpretation of slope

    -   Estimated ratio of rates for two subjects differing by 1 in their BMI

    -   Interpretation by exponentiation of slope

        -   A subject with a 1 $\textrm{kg} / m^2$ higher BMI will have an acid reflux event rate that is $2.3\%$ higher. (calc: $e^{0.0223} = 1.023$)

        -   We are 95% confident that the increase in event rate is between $1.3\%$ higher and $3.2\%$ higher

        -   There is a significant association between BMI and reflux events $p < 0.001$

## Example: Acid reflux and BMI by esophagitis status

### BMI modeled as a linear term

-   The following results compare using a Poisson model to a linear regression model

-   Both models will control for Esophagitis status, so any interpretation must involve "Holding esophagitis status constant..." ("Among subjects with the same Esophagitis status...")

-   Note the different (numerical) estimates for the coefficients and standard errors for BMI and esophagitis, but the similar statistical significance

-   Also if we plot the predicted number of events per day versus BMI, the results are similar from either model

#### R: Poisson regression of events with offset for log(total time monitored)

```{r}
m2.poisson <- glm(events ~ bmi + esop + offset(log(totalmins)), data=bmi.data, family="poisson")
m2.poisson

coeftest(m2.poisson, vcov=sandwich)
confint(coeftest(m2.poisson, vcov=sandwich))
```

#### R: Linear regression of rate (events/time) as outcome

```{r}
bmi.data$rate <- bmi.data$events / bmi.data$totalmins
m2.lm <-  lm(rate ~ bmi + esop, data=bmi.data)
m2.lm
coeftest(m2.lm, vcov=sandwich)
confint(coeftest(m2.lm, vcov=sandwich))
```

#### Stata Output

``` stata
. poisson events bmi esop, offset(logmins) robust

Poisson regression                                Number of obs   =        279
                                                  Wald chi2(2)    =      30.30
                                                  Prob > chi2     =     0.0000
Log pseudolikelihood = -11072.339                 Pseudo R2       =     0.0761

------------------------------------------------------------------------------
             |               Robust
      events |      Coef.   Std. Err.      z    P>|z|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
         bmi |   .0197465   .0047721     4.14   0.000     .0103934    .0290997
        esop |   .2622171    .083202     3.15   0.002     .0991442      .42529
       _cons |  -3.089033   .1423038   -21.71   0.000    -3.367944   -2.810123
     logmins |   (offset)
------------------------------------------------------------------------------


. gen eventsmins = events / mins
. regress eventsmins bmi esop, robust

Linear regression                                      Number of obs =     279
                                                       F(  2,   276) =   14.16
                                                       Prob > F      =  0.0000
                                                       R-squared     =  0.0856
                                                       Root MSE      =  .05102

------------------------------------------------------------------------------
             |               Robust
  eventsmins |      Coef.   Std. Err.      t    P>|t|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
         bmi |    .001839   .0004618     3.98   0.000     .0009299    .0027482
        esop |    .025104   .0085449     2.94   0.004     .0082826    .0419254
       _cons |   .0278461   .0129053     2.16   0.032     .0024407    .0532515
------------------------------------------------------------------------------
```

#### Comparison of predicted number of events from linear regression and Poisson regression models

-   Example prediction calculations: BMI=30, with esophagitis

    -   Linear regression: $0.0278461 + .025104 + .001839 \times 30 = 0.108$

        -   Stata: `adjust bmi=30 esop=1`

    -   Poisson regression: $e^{-3.089033 + 0.2622171 + .01975465 \times 30} = 0.107$

        -   Stata: `adjust bmi=30 esop=1, nooffset exp`

    -   Remember the above rates are for a 2-minute time interval. To convert to daily rates, multiply by 720

```{r}
# Predicted events per two minute time interval
predict(m2.lm, newdata=data.frame(bmi=30,esop=1))
exp(predict(m2.poisson, newdata=data.frame(bmi=30,esop=1,totalmins=1), type="link"))


# Predicted events per day (720 2-minute time intervals per day)
predict(m2.lm, newdata=data.frame(bmi=30,esop=1))*720
exp(predict(m2.poisson, newdata=data.frame(bmi=30,esop=1,totalmins=720), type="link"))
```

```{r}
bmi.data$esop.factor <- factor(bmi.data$esop, levels=0:1, labels=c("Esop Neg", "Esop Pos"))
m.spline2.adj <- glm(events ~ bmi + offset(log(totalmins)) + esop, data=bmi.data, family="poisson")
m.spline3.adj <- lm(events / totalmins ~ bmi + esop, data=bmi.data)
par(mfrow=c(1,2), mar=c(5,4,4,0.5))
plot(18:40, exp(predict(m.spline2.adj, newdata=data.frame(bmi=18:40, totalmins=720, esop=1), type="link")), type='l', ylab="Predicted number of events per day", xlab="BMI", ylim=c(30,100), main="Poisson Reg")
axis(4, labels=FALSE, ticks=TRUE)
legend("bottomright", c("Esophagitis Pos","Esophagitis Neg"), inset=0.05, col=1, lty=1:2)
lines(18:40, exp(predict(m.spline2.adj, newdata=data.frame(bmi=18:40, totalmins=720, esop=0), type="link")), lty=2)
plot(18:40, 720*predict(m.spline3.adj, newdata=data.frame(bmi=18:40, esop=1), type="response"), type='l', col='Blue', ylab="", xlab="BMI", ylim=c(30,100), main="Linear Reg", axes=FALSE)
axis(1)
axis(4)
axis(2, labels=FALSE, ticks=TRUE)
box()
lines(18:40, 720*predict(m.spline3.adj, newdata=data.frame(bmi=18:40, esop=0), type="response"), type='l', col='Blue', lty=2)
legend("bottomright", c("Esophagitis Pos","Esophagitis Neg"), inset=0.05, col="Blue", lty=1:2)
```

### BMI modeled using splines

-   Regression splines are handled more naturally in R than in Stata

    -   `glm(events ~ ns(bmi,4) + esop + offset(log(totalmins)), data=bmi.data, family="poisson")`

    -   `ns(bmi, 4)` specified a natural spline for bmi with 4 degrees of freedom

-   Note that there is an optical illusion in the following plots

    -   For both plots, it appears as if the lines are closer in the middle ranges of BMI

    -   For the Poisson regression, the true distance between lines is increasing with increasing with BMI

    -   For the Linear regrression, the true distance between lines is constant

```{r}
m.spline2.adj <- glm(events ~ ns(bmi,4) + offset(log(totalmins)) + esop, data=bmi.data, family="poisson")
m.spline3.adj <- lm(events / totalmins ~ ns(bmi,4) + esop, data=bmi.data)
par(mfrow=c(1,2), mar=c(5,4,4,0.5))
plot(18:40, exp(predict(m.spline2.adj, newdata=data.frame(bmi=18:40, totalmins=720, esop=1), type="link")), type='l', ylab="Predicted number of events per day", xlab="BMI", ylim=c(0,100), main="Poisson Reg")
axis(4, labels=FALSE, ticks=TRUE)
legend("bottomright", c("Esophagitis Pos","Esophagitis Neg"), inset=0.05, col=1, lty=1:2)
lines(18:40, exp(predict(m.spline2.adj, newdata=data.frame(bmi=18:40, totalmins=720, esop=0), type="link")), lty=2)
plot(18:40, 720*predict(m.spline3.adj, newdata=data.frame(bmi=18:40, esop=1), type="response"), type='l', col='Blue', ylab="", xlab="BMI", ylim=c(0,100), main="Linear Reg", axes=FALSE)
axis(1)
axis(4)
axis(2, labels=FALSE, ticks=TRUE)
box()
lines(18:40, 720*predict(m.spline3.adj, newdata=data.frame(bmi=18:40, esop=0), type="response"), type='l', col='Blue', lty=2)
legend("bottomright", c("Esophagitis Pos","Esophagitis Neg"), inset=0.05, col="Blue", lty=1:2)
```

### Comparison of modeling linear BMI to using spline function

-   For all regression models, we are more confident modeling associations than predicting means

-   When we use a linear term (i.e. a straight line) for the predictor, we are modeling a first-order association

    -   Most power to detect this type of association

    -   Always need to check that a first-order association answers the scientific question

        -   Counter example: Interested in seasonal trends in air pollution. A linear effect of time would only answer if air pollution levels are increasing/decreasing over time, not how they are changing from month to month

-   Flexible functions for predictors, including splines, are, in general, more useful if we care about predicting means or individual observations

-   Acid reflux example: Which model you choose depends on the scientific goals

    -   Primary goal: Is there an association between BMI and the rate of acid reflux?

        -   Fitting the linear BMI term answers this question

    -   Secondary goal: Describe the (mean) trend in reflux rates as a function of BMI

        -   A priori, I would be less inclined to believe a linear function captures the true mean relationship

        -   To answer this scientific question, a spline analysis is preferred

### Bayesian Poisson Regression

#### Likelihood Function

-   The Bayesian analysis begins by specifying the likelihood and prior distributions.

-   For the Poisson model, the likelihood function can bewritten as:

$$L(y | \lambda) = \prod_{i=1}^{n} \frac{e^{-\lambda_i} 
\lambda_i^{y_i}}{y_i!}$$

where $y_i$ is the observed count for each observation $i$, and $\lambda_i$ is the expected count.

-   In a Poisson regression model, $\lambda$ is modeled as a function of the covariates **X** and coefficients $\boldsymbol{\beta}$, such that:

$$\lambda = \exp(\textbf{X} \boldsymbol{\beta})$$

-   This equation links the expected counts to the linear predictor, allowing for the estimation of the effects of the covariates on the response variable.

-   (With the Poisson likelihood, we are still assuming the mean is equal to the variance)

#### Prior specification

-   As with any Bayesian analysis, prior distributions must be specified for all parameters in the model

-   For the Poisson regression model, $\boldsymbol{\theta} = \boldsymbol{\beta}$

-   A convenient choice of prior distribution for $\boldsymbol{\beta}$ is the Multivariable Normal distribution

$$
\pi(\boldsymbol{\beta}) \sim \mathcal{N}(\boldsymbol{\mu}, \boldsymbol{\Sigma})
$$

-   Here, $\boldsymbol{\mu}$ represents the mean vector of the prior distribution, and $\boldsymbol{\Sigma}$ represents the covariance matrix.

-   The prior mean $\boldsymbol{\mu}$ and covariance $\boldsymbol{\Sigma}$ can be chosen based on prior knowledge or elicited from experts, or can be set to non-informative values for a more objective analysis.

#### Posterior Distribution

-   The posterior distribution of $\boldsymbol{\beta}$ is proportional to the product of the likelihood function and the prior distribution:

$$
\pi(\boldsymbol{\beta} | \textbf{y}) \propto L(\textbf{y} | \lambda) 
\cdot f(\boldsymbol{\beta})
$$

-   Substituting the Poisson likelihood and multivariate normal prior, we get:

$$\pi(\boldsymbol{\beta} | \textbf{y}) \propto \prod_{i=1}^{n} 
\frac{e^{-\lambda_i} \lambda_i^{y_i}}{y_i!} \cdot \frac{1}{\sqrt{(2\pi)^p 
|\boldsymbol{\Sigma}|}} \exp\left(-\frac{1}{2} (\boldsymbol{\beta} - 
\boldsymbol{\mu})^T \boldsymbol{\Sigma}^{-1} (\boldsymbol{\beta} - 
\boldsymbol{\mu})\right)
$$

-   Where $\lambda_i = \exp(\textbf{x}_i \cdot \boldsymbol{\beta})$, and $p$ is the number of regression coefficients.

-   The posterior distribution is typically complex and does not have a closed-form expression, requiring numerical methods such as Markov chain Monte Carlo (MCMC) for inference.

    -   In this course, we have been using `rstanarm` and `stan_glm` to perform MCMC

### Bayesian Poisson regression example

* To fit a Bayesian Poisson regression model using the `rstanarm` package, we can use the `stan_glm()` function.

* The following code specifies a Bayesian model that is equivalent to the earlier frequentist model with bmi and esop and an offset

```{r}
library(rstanarm)

m2.bayes.poisson <- stan_glm(
  events ~ bmi + esop + offset(log(totalmins)),
  data = bmi.data,
  family = poisson(link = "log"),
  prior = normal(location = 0, scale = 1),
  prior_intercept = normal(location = 0, scale = 1),
  refresh=0,
  seed=12345
)
summary(m2.bayes.poisson, digits=4, prob=c(.025, .5, .975))
```

-   In this code:

    -   `events ~ bmi + esop + offset(log(totalmins))` specifies the model formula.
    -   `family = poisson(link = "log")` specifies a Poisson distribution with a log link function
    -   `prior = normal(location = 0, scale = 1)` and `prior_intercept = normal(location = 0, scale = 1)` specify priors for the regression coefficients and intercept.
    -   `seed = 1234` sets a random seed to ensure reproducibility of the results.
    -   `refresh = 0` suppresses the progress bar during model fitting.

-   And a summary of the prior distributions used

```{r}
prior_summary(m2.bayes.poisson, digits = 2)
```

#### Plot of Prior, Likelihood, Posterior

-   The following shows the prior, likelihood and the posterior for the Poisson model of the $\beta$ for BMI.

```{r}

plp.plotdata <- data.frame(post.draws=as.matrix(m2.bayes.poisson)[,"bmi"])

density_plot <- ggplot(plp.plotdata, aes(x = post.draws)) +
  geom_density(aes(fill = "Posterior (MCMC)"), color = NA, alpha = 0.5) +
  labs(x = "BMI", y = "Density") +
  scale_fill_manual(values = c("Posterior (MCMC)" = "blue")) +
  stat_function(fun = dnorm, args = list(mean = 0.0, sd = 1), aes(color = "Prior"), size=1) +
  scale_color_manual(name="Legend", values = c("Prior" = "red", "Likelihood (Asymptotic)" = "green")) +
  stat_function(fun = dnorm, args = list(mean = coef(m2.poisson)["bmi"], sd = 
sqrt(vcov(m2.poisson)["bmi","bmi"])), aes(color = "Likelihood (Asymptotic)"), size=1) +
  theme_minimal() + theme(legend.title=element_blank())

print(density_plot)
```

-   Here is just the posterior distributions for BMI and `esop`

```{r}
library(bayesplot)
mcmc_dens(m2.bayes.poisson, pars=c("bmi","esop")) +
  labs(title = "Posterior Distribution of bmi",
       subtitle = "Bayesian Poisson Regression Model",
       x = "Value", y = "Density") +
  theme_classic()
```

```{r}
 posterior_vs_prior(m2.bayes.poisson, pars=c("bmi","esop"))
```

* A comparison of the coefficients estimates from the frequentist and Bayesian approaches

```{r}
rbind(glm=coef(m2.poisson),
      Bayes.weak=coef(m2.bayes.poisson))
```

* And a comparsion of the estimated standard errors for the intercept and the slopes
```{r}
rbind(glm=sqrt(diag(vcov(m2.poisson))),
      glm.robust=sqrt(diag(sandwich(m2.poisson))),
      Bayes.weak=sqrt(diag(vcov(m2.bayes.poisson))))
```


* Note that the standard errors are quite different from the roubst approach

* From descriptive statistics, the mean and variance are quite different.  The standard Poisson model assumes the mean equals the variance.

```{r}
c(mean=mean(bmi.data$rate), var=var(bmi.data$rate))
library(dplyr)
bmi.data <- bmi.data %>% mutate(bmi_quartile = cut2(bmi, g = 4))

bmi.data %>% group_by(bmi_quartile, esop) %>% summarise(mean_rate = mean(rate, na.rm = TRUE), var_rate = var(rate, na.rm = TRUE))

```

