---
title: "Regression Based Prediction"
subtitle: "Lecture 13"
name: Lec13.Prediction.qmd
---

```{r}
library(rms)
library(ggplot2)
library(kableExtra)
library(pROC)
tryCatch(source('pander_registry.R'), error = function(e) invisible(e))
```

## Connection to Model Choice

-   In Lecture 12, we compared different approaches to modeling dose-response relationships (ANOVA, linear, polynomial, threshold models)

-   The choice of model directly affects the quality of predictions

    -   A model that is too simple (e.g., linear when the truth is nonlinear) will produce biased predictions for some groups

    -   A model that is too complex (e.g., ANOVA with many groups) will produce imprecise predictions

-   In this lecture, we shift focus from *which model to choose* to *how to use a chosen model for prediction*

    -   Prediction of summary measures (means, probabilities) with confidence intervals

    -   Prediction of individual observations with prediction intervals

    -   Classification and discrimination for binary outcomes

## Overview

### Uses of Regression

-   Other uses of regression beyond association and prediction

    -   Most of this course focuses on quantifying associations and making inferences

    -   Prediction is the focus of this lecture

    -   But regression-based thinking extends to many other areas, including:

        -   **Joint modeling**: Simultaneously modeling a longitudinal outcome and a time-to-event outcome

            -   e.g. Modeling repeated PSA measurements and time to cancer recurrence together

            -   Accounts for the dependence between the longitudinal trajectory and the event process

        -   **Latent variable and mixture models**: Identifying unobserved subgroups or constructs

            -   e.g. Latent class growth models to find distinct trajectories of disease progression

            -   Structural equation models relating observed measurements to latent constructs

        -   **Clustering and classification**: Dividing populations into subgroups based on patterns of covariates

            -   Regression trees, random forests, and other machine learning approaches

            -   "Small $n$, large $p$" problems (few subjects, many predictors)

        -   Getting valid inference in these settings is particularly challenging and is an area of active research in Biostatistics

-   Quantifying and comparing distributions

    -   Compare distribution of the outcome variable across levels of the grouping variable or variables

        -   Compare salary in males and females, across professors of the same rank, degree, experience, and field

        -   Compare FEV in smokers and non-smokers, across teenagers of the same age and height

    -   Need to decide which summary measure is appropriate for describing the distribution

        -   Common summary measures: Means, median, geometric mean, odds, rates, hazards

        -   Other measures: Variance, skewness, kurtosis, likelihood of extreme values, quantiles, etc.

    -   May desire estimates within specific subgroups

        -   Estimates of the association within gender, race, or age groups

        -   (This is effect modification)

-   Regression based prediction

    -   Prediction of summary measures

        -   Point prediction

            -   Best single estimate for the measurement that would be obtained, on average, for many individuals with given covariates

        -   Interval prediction

            -   Quantify the uncertainty of the average

            -   Range of summary measure that might be reasonable to observe

    -   Prediction of individual measurements

        -   Point prediction

            -   Best single estimate for the measurement that would be obtained on a single individual with given covariates

        -   Interval prediction

            -   Quantify the uncertainty of the individual prediction

            -   Range of measurements that might reasonably be observed in a future individual with given covariates

### Examples of Prediction

#### Continuous prediction: Creatinine clearance

-   Creatinine is a continuously produced breakdown product in muscles

-   Removed by the kidney through filtration

-   Amount of creatinine cleared by the kidneys in 24 hours is a measure of renal function

    -   Gold standard is to collect urine output for 24 hours, measure creatinine

    -   Would prefer to find a combination of blood and urine measures that can be obtained at one time point, yet still provides accurate prediction of a patient's creatinine clearance

-   Statistical approach

    -   Collect a training dataset to build a regression model for prediction

        -   Measure true creatinine clearance

        -   Measure age, gender, weight, height, blood markers, urinary markers, etc.

        -   Fit a (data driven?) regression model to predict true creatinine clearance

    -   Collect a validation dataset

        -   Use the regression estimates (the $\beta$s) from the training dataset to see how well your model predict creatinine clearance

        -   Quantify the accuracy of the predictive model (e.g. mean squared error)

        -   Cross validation

#### Discrimination (binary prediction): Low birth weight

-   Want to predict which infants are more likely to be less than 2500 grams at birth

-   Possible predictors: Age, race, blood biomarkers collected during pregnancy

-   Statistical Approach

    -   Collect a training dataset to build a regression model for prediction

        -   Measure birth weight

        -   Measure age, race, blood biomarkers

        -   Fit a (data driven?) regression model to predict low birth weight

    -   Collect a validation dataset

        -   Use the regression estimates (the $\beta$s) from the training dataset to see how well your model predict low birth weight

        -   Quantify the accuracy of the predictive model

            -   Sensitivity, specificity (ROC curves)

            -   Predictive value positive, predictive value negative

#### Interval prediction: Normal ranges of PSA

-   Identify the range of PSA values that would be expected in 95% of most healthy adult males

-   Possibly stratify by age, race

-   Statistical Approach

    -   Collect a training dataset to build a regression model for prediction

        -   Measure PSA and variable to predict PSA

        -   Need to estimate the quantiles

            -   Mean plus/minus 2 standard deviations (makes strong Normality assumption)

            -   Estimate the quantiles, provide CIs around the quantiles (likely low precision)

    -   Collect a validation dataset

        -   Quantify the accuracy of the predictive model (e.g. coverage probabilities)

### Assumptions needed for inference

#### Inference for associations

-   Necessary assumptions for classical regression (no robust standard errors)

    -   Independence of response measurements within identified clusters

    -   Have appropriately modeled the within group variance

        -   Linear regression: Equal variance across groups

        -   Other regressions: Appropriate mean-variance relationship

            -   Lack of model fit may lead to poor estimate of the variance

    -   Sample size is large enough so that parameter estimates approximately follow a Normal distribution

-   Necessary assumptions for first order trends using robust standard errors

    -   Independence of response measurements within identified clusters

    -   (Robust standard errors accounts for heteroscedasticity in large samples)

        -   Lack of model fit leads to conservative inference due to mixing systematic and random error

    -   Sample size is large enough so that parameter estimates approximately follow a Normal distribution

#### Inference for predictions

-   Additional assumptions for predictions of means

    -   Our regression model has accurately described the relationship between summary measures across groups

    -   For continuous covariates, often involves flexible models that allow for departures from linearity

-   Additional assumptions for predictions of individual observations

    -   Also need to know the shape of the distribution within each group

    -   Methods implemented in software often rely on strong assumptions like Normality

## Estimation (prediction) of summary measures

-   Given age, height, and sex, estimate the mean (or geometric mean) FEV

    -   Use linear regression to obtain estimates and CI

-   Given age and PSA, estimate the probability (or odds) of remaining in remission for 24 months

    -   Use logistic regression to obtain estimates and CI


-   Assumptions are the same as for inference for predictions discussed above, with the key addition that the regression model must accurately describe the relationship across groups

### Point estimates

-   Point estimates obtained by substitution of predictor values into the estimated regression equation

    -   For linear regression: substitute covariate values and compute the fitted value directly

    -   For logistic regression: compute the linear predictor (log-odds), then transform to odds or probability

    -   We demonstrate both below with the FEV data

### Interval estimates

-   If assumptions hold, interval estimates can be obtained

    -   We generally find a confidence interval for the transformed quantity, and then back transform to the desired quantity

        -   For logistic regression, calculate the CI for the log odds, then transform to odds or probability

-   Statistical criteria for determining the "best" estimate

    -   Consistent: Correct estimate (with infinite sample size)

    -   Precise: Minimum variance among (unbiased) estimators

    -   Common regression methods provide the best estimate in a wide variety of settings

-   Necessary assumptions

    -   Independence

    -   Variance approximated by the model (relaxed for robust SE)

    -   **Regression model accurately describes the relationship among summary measures across groups**

    -   Sufficient sample size so that asymptotic Normality of the parameter estimates holds

-   Interval estimates

    -   When we substitute in the predictor values, it provides an estimate of the model transformation of the summary measure

    -   Model transformation of the summary measure varies by regression setting

        -   Linear regression: Mean

        -   Linear regression on log transformed outcome: Log geometric mean

        -   Logistic regression: Log odds

        -   Poisson regression: Log rate

    -   Formulas for the confidence interval

        -   In general: $\textrm{(estimate)} \pm \textrm{(crit value)} \times \textrm{(std error)}$

        -   In linear regression, the $t$ distribution is usually used to obtain the confidence interval

            -   $\textrm{(crit value)} = \texttt{qt}(1-\alpha/2, df)$

            -   Degrees of freedom $df = n - \textrm{number of predictors in model}$

        -   In other regressions, we use the standard Normal distribution

            -   $\textrm{(crit value)} = \texttt{qnorm}(1-\alpha/2) = 1.96$

-   Interval estimates in R

    -   After storing a model, the R function `predict` will compute estimates and standard errors

    -   `predict(object, se.fit=FALSE, ...)`

        -   `object` is the stored fitted model

        -   `se.fit` will provide the standard error of the fit (defaults to `FALSE`)

        -   See `help(predict)` for more details and options

### Example: Predicting geometric mean FEV and odds of smoking

-   This is the **continuous prediction** case (analogous to the creatinine clearance example) and the **binary discrimination** case (analogous to the low birth weight example) from the overview

```{r}
fev <- read.csv(file="data/FEV.csv")
fev$logfev <- log(fev$fev)
fev$smoker <- (fev$smoke == "current smoker") + 0
```

-   Linear regression of log(FEV) on height and age

```{r}
m.fev <- lm(logfev ~ height + age, data=fev)
summary(m.fev)
```

-   Predict geometric mean FEV for a subject with age=10 and height=66

```{r}
newdata <- data.frame(age=10, height=66)
pred.fit <- predict(m.fev, newdata=newdata, se.fit=TRUE)
crit <- qt(0.975, df=m.fev$df.residual)

# Geometric mean and CI (back-transform from log scale)
gmfev <- exp(pred.fit$fit)
gm.ci <- exp(pred.fit$fit + c(-1, 1) * crit * pred.fit$se.fit)

data.frame(gmfev = gmfev, gm.lo = gm.ci[1], gm.hi = gm.ci[2]) |>
  kable(digits=4, col.names=c("Geometric Mean FEV", "95% CI Lower", "95% CI Upper"),
        caption="Predicted geometric mean FEV (age=10, height=66)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   Logistic regression of smoking on height and age

```{r}
m.smoke <- glm(smoker ~ height + age, data=fev, family=binomial)
summary(m.smoke)
```

-   Predict odds and probability of smoking for a subject with age=10 and height=66

```{r}
pred.logodds <- predict(m.smoke, newdata=newdata, se.fit=TRUE)

# Odds and CI
odds <- exp(pred.logodds$fit)
odds.ci <- exp(pred.logodds$fit + c(-1, 1) * 1.96 * pred.logodds$se.fit)

# Probability and CI (back-transform from log-odds)
prob <- plogis(pred.logodds$fit)
prob.ci <- plogis(pred.logodds$fit + c(-1, 1) * 1.96 * pred.logodds$se.fit)

data.frame(
  Odds = odds, Odds.Lo = odds.ci[1], Odds.Hi = odds.ci[2],
  Prob = prob, Prob.Lo = prob.ci[1], Prob.Hi = prob.ci[2]) |>
  kable(digits=4,
        col.names=c("Odds", "95% CI Lower", "95% CI Upper",
                     "Probability", "95% CI Lower", "95% CI Upper"),
        caption="Predicted odds and probability of smoking (age=10, height=66)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

## Prediction (forecasting) of individual observations

-   Methods are for continuous outcomes

    -   Will deal with discrimination of binary outcomes in next section

-   Given age, height, and sex, predict a new subject's FEV

    -   Use linear regression to obtain estimates and prediction interval

### Assumptions

-   Predicting individual observations requires **stronger assumptions** than predicting summary measures

    -   All the assumptions for predicting means, plus:

    -   Variance approximated by the model — **NOT** relaxed for robust SE

    -   Shape of the distribution must be the same in each group (e.g., Normality)

        -   Transformation of the outcome may be very useful

-   These assumptions are very strong

    -   Consequently, we do not have many methods that provide robust inference

        -   In general, I prefer methods that make as few assumptions as possible

        -   Robust standard errors will not help in this situation

    -   Proper transformation of outcomes and predictors may be necessary so that underlying assumptions of classical linear regression model hold

        -   Models, estimates, will need to be appropriately penalized for valid inference

        -   Topic of more advanced regression courses (Regression Modeling Strategies)

    -   Precise methods have been developed for

        -   Binary variables (the mean specifies the variances)

        -   Continuous data that follow a Normal distribution

### Point and interval estimates for Normal data

-   Point estimates are obtained by the same substitution as for summary measures — but the interpretation changes: we are now forecasting a single individual's value, not a group mean

-   Interval estimates

    -   Under appropriate assumptions, we can obtain standard errors for such predictions

    -   Standard errors must account for two sources of variability

        -   Variability in estimating the regression parameters (same as in predictions of summary measures)

        -   Variability due to subject

            -   Additional variability about the sample mean; "within group standard deviation"

            -   Estimating this source of variability is where the additional Normality assumption is key

    -   Formulas for the prediction interval

        -   In general: $\textrm{(prediction)} \pm \textrm{(crit value)} \times \textrm{(std error)}$

        -   In linear regression, the $t$ distribution is usually used to obtain the prediction interval

            -   $\textrm{(crit value)} = \texttt{qt}(1-\alpha/2, df)$

            -   Degrees of freedom $df = n - \textrm{number of predictors in model}$

-   Interval estimates in R

    -   After storing a model, `predict` provides estimates and standard errors of the fit

    -   For the standard error of the forecast (individual prediction), we need to include the root mean squared error

        -   $\textrm{se}_\textrm{pred}^2 = \textrm{se}_\textrm{fit}^2 + \textrm{RMSE}^2$

    -   Alternatively, use `predict(object, newdata, interval="prediction")` which computes prediction intervals directly

-   General comment about software

    -   Software typically implements prediction intervals by assuming Normal data

    -   Carefully investigate and understand how any software or package is making predictions

        -   Don't trust the "black box" to be giving you results that are widely applicable to all situations

    -   Be careful that you understand exactly what is going on if you are interested in making predictions

### Example: Prediction interval for FEV

-   This is the **interval prediction** case (analogous to establishing normal ranges of PSA from the overview)

-   Using the same FEV model, compare the confidence interval for the mean with the prediction interval for an individual

```{r}
# Confidence interval for the mean (summary measure)
ci.mean <- predict(m.fev, newdata=newdata, interval="confidence")

# Prediction interval for an individual observation
pi.indiv <- predict(m.fev, newdata=newdata, interval="prediction")

# Back-transform to FEV scale (geometric mean)
comparison <- data.frame(
  Type = c("Summary Measure (CI)", "Individual Prediction (PI)"),
  Estimate = exp(c(ci.mean[,"fit"], pi.indiv[,"fit"])),
  Lower = exp(c(ci.mean[,"lwr"], pi.indiv[,"lwr"])),
  Upper = exp(c(ci.mean[,"upr"], pi.indiv[,"upr"]))
)

kable(comparison, digits=4,
      col.names = c("Type", "Geometric Mean FEV", "Lower", "Upper"),
      caption="Comparing CI for mean vs prediction interval (age=10, height=66)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   Point estimates are identical

-   The prediction interval for individual observations is much wider

-   As $n$ increases, the width of the prediction interval (on the log scale) approaches $\pm 1.96 \times \textrm{RMSE}$

-   As $n$ increases, the width of the interval around the summary measure approaches $0$

-   We can verify the relationship between the standard errors

```{r}
# Demonstrate: se_pred^2 = se_fit^2 + RMSE^2
pred.fit <- predict(m.fev, newdata=newdata, se.fit=TRUE)
rmse <- summary(m.fev)$sigma

se.fit <- pred.fit$se.fit
se.pred <- sqrt(se.fit^2 + rmse^2)

data.frame(
  se.fit = se.fit, se.fit.sq = se.fit^2,
  RMSE = rmse, RMSE.sq = rmse^2,
  se.pred = se.pred, se.pred.sq = se.pred^2
) |>
  kable(digits=7, col.names=c("SE(fit)", "SE(fit)^2", "RMSE", "RMSE^2", "SE(pred)", "SE(pred)^2"),
        caption = "Decomposition of prediction standard error") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   $\textrm{se}_\textrm{pred}^2 = \textrm{se}_\textrm{fit}^2 + \textrm{RMSE}^2$

-   In R, `predict(..., interval="prediction")` handles this calculation automatically

### Mathematics of Prediction Intervals

#### Recap: Simple linear regression (single predictor)

-   Model: $Y_i | X_i \sim N(\beta_0 + \beta_1 X_i, \sigma^2)$

-   Alternative specification: $Y_i | X_i = \beta_0 + \beta_1 X_i + \epsilon_i$, where $\epsilon_i \sim N(0, \sigma^2)$

-   Estimated mean: $\hat{\beta}_0 + \hat{\beta}_1 X_h \sim N\left(\beta_0 + \beta_1 X_h, \sigma^2 V\right)$

-   Predicted value of individual observation: $\hat{\beta}_0 + \hat{\beta}_1 X_h + \epsilon_h \sim N\left(\beta_0 + \beta_1 X_h, \sigma^2 (1 + V)\right)$

    -   $V = \frac{1}{n} + \frac{(X_h - \overline{X})^2}{\sum_{i=1}^n (X_i - \overline{X})^2}$

    -   $X_h$ is the chosen value of the covariate (e.g. age==10)

    -   Note: As $n \rightarrow \infty, V \rightarrow 0$

#### Multivariable model (matrix notation)

-   For $p$ predictors, the linear regression model in matrix form is

$$\mathbf{Y} = \mathbf{X} \boldsymbol{\beta} + \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim N(\mathbf{0}, \sigma^2 \mathbf{I})$$

where $\mathbf{Y}$ is the $n \times 1$ response vector, $\mathbf{X}$ is the $n \times (p+1)$ design matrix (including intercept column), and $\boldsymbol{\beta}$ is the $(p+1) \times 1$ parameter vector

-   The OLS estimator is $\hat{\boldsymbol{\beta}} = (\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}$

-   For a new observation with covariate vector $\mathbf{x}_h$ (a $1 \times (p+1)$ row vector including a 1 for the intercept):

-   **Confidence interval for the mean response** (prediction of summary measure)

    -   $\hat{y}_h = \mathbf{x}_h \hat{\boldsymbol{\beta}}$

    -   $\textrm{Var}(\hat{y}_h) = \sigma^2 \mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h'$

    -   $\textrm{se}_\textrm{fit} = \hat{\sigma} \sqrt{\mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h'}$

    -   CI: $\hat{y}_h \pm t_{n-p-1, \alpha/2} \times \textrm{se}_\textrm{fit}$

-   **Prediction interval for a new individual observation**

    -   The new observation is $Y_h = \mathbf{x}_h \boldsymbol{\beta} + \epsilon_h$

    -   $\textrm{Var}(\hat{y}_h - Y_h) = \sigma^2 \left(1 + \mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h'\right)$

    -   $\textrm{se}_\textrm{pred} = \hat{\sigma} \sqrt{1 + \mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h'}$

    -   PI: $\hat{y}_h \pm t_{n-p-1, \alpha/2} \times \textrm{se}_\textrm{pred}$

-   Note: $\textrm{se}_\textrm{pred}^2 = \textrm{se}_\textrm{fit}^2 + \hat{\sigma}^2$, exactly the decomposition we demonstrated earlier

-   Note: As $n \rightarrow \infty$, $\mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h' \rightarrow 0$ and the prediction interval width approaches $\pm t \times \hat{\sigma}$ (the irreducible within-group variability)

#### Verification using matrix algebra in R

-   We can carry out these calculations by hand using the FEV model and verify they match `predict()`

```{r}
# Design matrix and estimated coefficients from the log(FEV) model
X <- model.matrix(m.fev)
beta.hat <- coef(m.fev)
sigma.hat <- summary(m.fev)$sigma
n <- nrow(X)
p <- ncol(X) - 1  # number of predictors (excluding intercept)

# New observation: age=10, height=66
xh <- matrix(c(1, 66, 10), nrow=1)  # [intercept, height, age]
colnames(xh) <- names(beta.hat)
```

```{r}
# Compute (X'X)^{-1}
XtX.inv <- solve(t(X) %*% X)

# Fitted value
yhat <- as.numeric(xh %*% beta.hat)

# Variance of the fitted value: sigma^2 * xh (X'X)^{-1} xh'
V <- as.numeric(xh %*% XtX.inv %*% t(xh))
se.fit.manual <- sigma.hat * sqrt(V)

# Variance of prediction: sigma^2 * (1 + xh (X'X)^{-1} xh')
se.pred.manual <- sigma.hat * sqrt(1 + V)

# Critical value
crit <- qt(0.975, df = n - p - 1)

# Confidence interval for mean (on log scale)
ci.manual <- yhat + c(-1, 1) * crit * se.fit.manual

# Prediction interval for individual (on log scale)
pi.manual <- yhat + c(-1, 1) * crit * se.pred.manual
```

```{r}
# Compare to predict() output
pred.ci <- predict(m.fev, newdata=newdata, interval="confidence")
pred.pi <- predict(m.fev, newdata=newdata, interval="prediction")
pred.se <- predict(m.fev, newdata=newdata, se.fit=TRUE)

comparison <- data.frame(
  Quantity = c("Fitted value", "SE(fit)", "SE(pred)",
               "CI lower", "CI upper", "PI lower", "PI upper"),
  `Matrix Algebra` = c(yhat, se.fit.manual, se.pred.manual,
                        ci.manual[1], ci.manual[2],
                        pi.manual[1], pi.manual[2]),
  `predict()` = c(pred.ci[,"fit"], pred.se$se.fit,
                   sqrt(pred.se$se.fit^2 + sigma.hat^2),
                   pred.ci[,"lwr"], pred.ci[,"upr"],
                   pred.pi[,"lwr"], pred.pi[,"upr"]),
  check.names = FALSE
)

kable(comparison, digits=6,
      caption = "Verification: matrix algebra vs predict() (log FEV scale, age=10, height=66)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   The matrix algebra calculations match `predict()` exactly

-   This confirms that `predict()` is computing $\mathbf{x}_h (\mathbf{X}'\mathbf{X})^{-1} \mathbf{x}_h'$ internally

-   Back-transforming to the FEV scale (exponentiating), these correspond to the geometric mean CI and prediction interval we computed earlier

## Prediction in the Beta Carotene Trial

-   In Lecture 12, we compared several models for the dose-response relationship in the beta carotene trial

-   Now we examine how model choice affects the predictions themselves, and how adjusting for baseline affects precision

### How model choice affects predictions

-   We refit the models from Lecture 12 and compare their predicted means and confidence intervals at each dose level

```{r}
carot <- stata.get("data/carot.dta")
carot$trt <- (carot$dose > 0) + 0
dose.levels <- c(0, 15, 30, 45, 60)
newdose <- data.frame(dose = dose.levels, carot0 = mean(carot$carot0),
                       trt = (dose.levels > 0) + 0)

# Fit the models (all adjusted for baseline)
m.anova  <- lm(carot3 ~ factor(dose) + carot0, data = carot)
m.linear <- lm(carot3 ~ dose + carot0, data = carot)
m.quad   <- lm(carot3 ~ poly(dose, 2) + carot0, data = carot)
m.dichot <- lm(carot3 ~ trt + carot0, data = carot)
m.thres  <- lm(carot3 ~ trt + dose + carot0, data = carot)
```

```{r}
# Predicted means and CIs from each model
get_preds <- function(model, nd) {
  p <- predict(model, newdata = nd, interval = "confidence")
  data.frame(dose = nd$dose, fit = p[,"fit"], lwr = p[,"lwr"], upr = p[,"upr"])
}

pred.anova  <- get_preds(m.anova,  newdose)
pred.linear <- get_preds(m.linear, newdose)
pred.quad   <- get_preds(m.quad,   newdose)
pred.dichot <- get_preds(m.dichot, newdose)
pred.thres  <- get_preds(m.thres,  newdose)
```

```{r, fig.width=10, fig.height=6, fig.cap="Predicted mean plasma beta-carotene (with 95% CI) by dose, for five different models. All models adjust for baseline and predictions are at the mean baseline value."}
# Colors and labels for each model
cols <- c("black", "blue", "red", "darkgreen", "purple")
names <- c("ANOVA", "Linear", "Quadratic", "Dichotomized", "Threshold + Linear")
preds <- list(pred.anova, pred.linear, pred.quad, pred.dichot, pred.thres)
offsets <- c(-1.2, -0.6, 0, 0.6, 1.2)

plot(NULL, xlim = c(-3, 63), ylim = c(-200, 2800),
     xlab = "Dose (mg/day)", ylab = "Predicted Plasma Beta-Carotene",
     main = "Model Choice Affects Predictions")

for(i in 1:5) {
  d <- preds[[i]]
  x <- d$dose + offsets[i]
  arrows(x, d$lwr, x, d$upr, angle = 90, code = 3, length = 0.04,
         col = cols[i], lwd = 1.5)
  points(x, d$fit, pch = 19, col = cols[i], cex = 1.2)
  lines(d$dose + offsets[i], d$fit, col = cols[i], lwd = 1.5, lty = i)
}

legend("topleft", names, col = cols, pch = 19, lwd = 1.5, lty = 1:5,
       bty = "n", cex = 0.9)
```

-   Key observations:

    -   The ANOVA model fits each dose group independently — widest CIs, but no risk of misspecifying the dose-response shape

    -   The linear model borrows strength across dose groups — narrower CIs, but the predictions are biased if the true relationship is not linear

    -   The more parsimonious models (dichotomized, linear) have narrower CIs at each dose level, but may miss important features of the dose-response

    -   The threshold + linear model captures both the jump from placebo to active and the linear trend among active doses — a good compromise

```{r}
# Compare CI widths across models at each dose level
ci_widths <- data.frame(
  Dose = dose.levels,
  ANOVA = pred.anova$upr - pred.anova$lwr,
  Linear = pred.linear$upr - pred.linear$lwr,
  Quadratic = pred.quad$upr - pred.quad$lwr,
  Dichotomized = pred.dichot$upr - pred.dichot$lwr,
  `Threshold + Linear` = pred.thres$upr - pred.thres$lwr,
  check.names = FALSE
)
kable(ci_widths, digits = 0,
      caption = "95% CI width by model and dose level (at mean baseline)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Adjusting for baseline: conditional and marginal predictions

-   In the beta carotene trial, we can form predictions three ways:

    1.  **Unadjusted group means**: The sample mean at each dose level, ignoring baseline entirely (marginal over baseline)

    2.  **ANCOVA conditional predictions**: Predicted 9-month level for a patient at a *specific* baseline value

    3.  **ANCOVA marginal means**: Predicted 9-month level at each dose, averaged over the observed distribution of baseline values (marginal over baseline, but leveraging the baseline-outcome correlation for precision)

-   Because this is a randomized trial, baseline should be balanced across dose groups

```{r}
# Descriptive: baseline beta-carotene by dose group
baseline_by_dose <- aggregate(carot0 ~ dose, data = carot,
  FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
baseline_df <- data.frame(Dose = baseline_by_dose$dose, baseline_by_dose$carot0,
                           check.names = FALSE)
kable(baseline_df, digits = 1,
      col.names = c("Dose", "N", "Mean Baseline", "SD Baseline"),
      caption = "Baseline plasma beta-carotene by dose group") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   Baseline means are similar across groups, as expected from randomization

-   We will use the threshold + linear model from Lecture 12 as our preferred model

```{r}
# 1. Unadjusted group means
m.unadj <- lm(carot3 ~ factor(dose), data = carot)
pred.unadj <- predict(m.unadj, newdata = data.frame(dose = dose.levels),
                       se.fit = TRUE)
crit.unadj <- qt(0.975, m.unadj$df.residual)

unadj <- data.frame(
  dose = dose.levels,
  fit = pred.unadj$fit,
  lwr = pred.unadj$fit - crit.unadj * pred.unadj$se.fit,
  upr = pred.unadj$fit + crit.unadj * pred.unadj$se.fit,
  type = "Unadjusted"
)
```

```{r}
# 2. ANCOVA conditional predictions at specific baseline values
crit.thres <- qt(0.975, m.thres$df.residual)

# Low baseline (25th percentile) and high baseline (75th percentile)
q25 <- quantile(carot$carot0, 0.25)
q75 <- quantile(carot$carot0, 0.75)

nd.low  <- data.frame(dose = dose.levels, carot0 = q25,
                       trt = (dose.levels > 0) + 0)
nd.high <- data.frame(dose = dose.levels, carot0 = q75,
                       trt = (dose.levels > 0) + 0)

pred.low  <- predict(m.thres, newdata = nd.low,  interval = "confidence")
pred.high <- predict(m.thres, newdata = nd.high, interval = "confidence")

cond.low <- data.frame(dose = dose.levels, pred.low,
                        type = sprintf("Conditional (baseline = %.0f)", q25),
                        check.names = FALSE)
cond.high <- data.frame(dose = dose.levels, pred.high,
                         type = sprintf("Conditional (baseline = %.0f)", q75),
                         check.names = FALSE)
```

```{r}
# 3. ANCOVA marginal means: average predictions over observed baseline distribution
# For each dose level, predict for every subject using their actual baseline,
# then average across subjects
marginal <- do.call(rbind, lapply(dose.levels, function(d) {
  nd <- data.frame(dose = d, carot0 = carot$carot0,
                   trt = (d > 0) + 0)
  preds <- predict(m.thres, newdata = nd)
  data.frame(dose = d, fit = mean(preds))
}))
marginal$type <- "ANCOVA Marginal"
```

-   Computing confidence intervals for marginal means requires methods beyond classical regression formulas, since the marginal mean is an average of predictions across the baseline distribution. We will return to this using the **bootstrap** in the bootstrap lecture notes.

```{r, fig.width=10, fig.height=6, fig.cap="Three approaches to estimating the dose-response: unadjusted means (with CIs), ANCOVA conditional predictions at low and high baseline (with CIs), and ANCOVA marginal means (point estimates only)"}
plot(NULL, xlim = c(-3, 63), ylim = c(-200, 2800),
     xlab = "Dose (mg/day)", ylab = "Plasma Beta-Carotene at 9 Months",
     main = "Unadjusted, Conditional, and Marginal Predictions")

# Unadjusted
arrows(unadj$dose - 1, unadj$lwr, unadj$dose - 1, unadj$upr,
       angle = 90, code = 3, length = 0.04, col = "gray50", lwd = 2)
points(unadj$dose - 1, unadj$fit, pch = 19, col = "gray50", cex = 1.5)

# Conditional: low baseline
arrows(cond.low$dose - 0.3, cond.low$lwr, cond.low$dose - 0.3, cond.low$upr,
       angle = 90, code = 3, length = 0.04, col = "steelblue", lwd = 2)
points(cond.low$dose - 0.3, cond.low$fit, pch = 17, col = "steelblue", cex = 1.3)

# Conditional: high baseline
arrows(cond.high$dose + 0.3, cond.high$lwr, cond.high$dose + 0.3, cond.high$upr,
       angle = 90, code = 3, length = 0.04, col = "coral", lwd = 2)
points(cond.high$dose + 0.3, cond.high$fit, pch = 17, col = "coral", cex = 1.3)

# Marginal (point estimates only — CIs require bootstrap, covered later)
points(marginal$dose + 1, marginal$fit, pch = 15, col = "darkgreen", cex = 1.5)

legend("topleft",
  c("Unadjusted group means",
    sprintf("Conditional (baseline = %.0f, Q1)", q25),
    sprintf("Conditional (baseline = %.0f, Q3)", q75),
    "ANCOVA marginal means (point est.)"),
  col = c("gray50", "steelblue", "coral", "darkgreen"),
  pch = c(19, 17, 17, 15), lwd = 2, bty = "n", cex = 0.9)
```

-   Key observations:

    -   The **unadjusted means** and **ANCOVA marginal means** estimate the same quantity — the population-average plasma level at each dose — and are very similar because randomization balanced baseline across groups

    -   The **conditional predictions** shift vertically depending on the baseline value: patients with higher baseline beta-carotene have higher predicted 9-month levels

    -   The conditional predictions at any specific baseline value have narrower CIs than the unadjusted means — this is the precision gain from ANCOVA

### Precision gain from ANCOVA

-   The precision gain from ANCOVA depends on the correlation between baseline and 9-month plasma levels ($\rho$)

    -   Recall from Lecture 12: the variance of the ANCOVA estimator is proportional to $\sigma^2(1 - \rho^2)/n$, compared to $\sigma^2/n$ for the unadjusted estimator

    -   Stronger baseline-outcome correlation means greater precision gain from adjustment

```{r}
rho <- cor(carot$carot0, carot$carot3)
sprintf("Correlation between baseline and 9-month plasma levels: %.3f", rho)
sprintf("Theoretical variance reduction factor: 1 - rho^2 = %.3f", 1 - rho^2)
```

-   We will quantify the precision gain more precisely in the bootstrap lecture notes, where we can compute CIs for both the unadjusted and marginal means and compare their widths directly

## Prediction of Binary Measurements

### Classification (discrimination)

-   Sometimes the scientific question is one of deriving a rule to classify subjects

    -   Diagnosis of prostate cancer: Based on age, race, and PSA, should we make a diagnosis of prostate cancer?

    -   Prognosis of patients with primary biliary cirrhosis: Based on age, bilirubin, albumin, edema, protime, is the patient likely to die within the next year?

-   Classification can be regarded as trying to predict the value of a binary variable

    -   Earlier, we were estimating the probability and odds of relapse within a particular group (a summary measure)

    -   Now we want to decide whether a particular individual will relapse or not (an individual measure)

    -   There is an obvious connection between the above two ideas

        -   The probability or odds tells us everything about the distribution of values

        -   The only possible values are 0 or 1

-   Typical approach

    -   First, use regression model to estimate the probability of event in each group

    -   Second, form a decision rule based on estimated probability of event

        -   If estimate $\geq c$ (or $\leq c$), predict outcome is 1

        -   If estimate $< c$ (or $> c$), predict outcome is 0

    -   Quantify the accuracy of the decision rule

    -   For disease $D$ and test $T$

        -   Sensitivity: $\textrm{Pr}(T^+ | D^+)$

        -   Specificity: $\textrm{Pr}(T^- | D^-)$

        -   Predicted Value Positive: $\textrm{Pr}(D^+ | T^+)$

        -   Predicted Value Negative: $\textrm{Pr}(D^- | T^-)$

### ROC Curves

-   Receiver operating characteristic (ROC) curves plot the sensitivity against the false positive rate (e.g. 1 - specificity)

-   Ideal tests will have both high sensitivity and high specificity (low false positives)

    -   Graphically depicted as points in the upper left portion of the plot

    -   1:1 line represents a test that is no better than the flip of a coin

-   Test usually involves the classification of some binary outcome by a continuous predictor (or set of continuous predictors)

    -   When the continuous predictor is above some point $c$, the test is said to be positive

    -   ROC curves plots the sensitivity and false positive rate for all possible values of $c$

-   Major drawbacks to ROC analysis

    -   Tempts analyst to find a cutoff point ($c$) when none really exists

    -   Better to treat your covariate or linear prediction as a continuous variable

### Visualizing discrimination and the ROC curve

-   To build intuition about the ROC curve, consider the relationship between a cutpoint on the risk score and the corresponding point on the ROC curve

    -   **Left panel**: Box plots of predicted risk scores, stratified by the actual binary outcome. A horizontal line shows the current cutpoint $c$

    -   **Right panel**: The ROC curve, with the point corresponding to that cutpoint highlighted

    -   As we lower the cutpoint, we classify more subjects as positive — sensitivity increases, but so does the false positive rate

```{r}
# Use the age-based smoking model for this demonstration
m.demo <- glm(smoker ~ age, data=fev, family=binomial)
fev$risk.score <- predict(m.demo, type="response")

# Compute the full ROC curve
roc.demo <- roc(fev$smoker, fev$risk.score, quiet=TRUE)

# Helper function for paired cutpoint display
plot_cutpoint <- function(cp, risk.score, outcome, roc.obj) {
  pred.pos <- risk.score >= cp
  sens <- sum(pred.pos & outcome == 1) / sum(outcome == 1)
  spec <- sum(!pred.pos & outcome == 0) / sum(outcome == 0)
  fpr <- 1 - spec

  par(mfrow=c(1, 2), mar=c(4, 4, 3, 1))

  # Left panel: Box plots with cutpoint line
  boxplot(risk.score ~ outcome, horizontal=FALSE,
          names=c("Non-smoker (D-)", "Smoker (D+)"),
          ylab="Predicted Risk Score", col=c("lightblue","lightsalmon"),
          main=sprintf("Cutpoint c = %.2f", cp), ylim=c(0,1))
  abline(h=cp, col="red", lwd=2.5, lty=2)
  text(0.6, cp + 0.04, sprintf("c = %.2f", cp), col="red", cex=1.1, font=2)

  # Right panel: ROC curve with highlighted point
  plot(1 - roc.obj$specificities, roc.obj$sensitivities,
       type="l", lwd=2, col="gray40",
       xlab="False Positive Rate (1 - Specificity)",
       ylab="Sensitivity",
       main=sprintf("Sens = %.0f%%    Spec = %.0f%%", 100*sens, 100*spec),
       xlim=c(0,1), ylim=c(0,1))
  abline(0, 1, col="gray70", lty=3)
  points(fpr, sens, col="red", pch=19, cex=3)
  text(fpr + 0.10, sens - 0.07,
       sprintf("(%.2f, %.2f)", fpr, sens), col="red", cex=1.1, font=2)
}
```

::: {.panel-tabset}

#### c = 0.70

```{r, fig.height=5, fig.width=10}
plot_cutpoint(0.70, fev$risk.score, fev$smoker, roc.demo)
```

Very high cutpoint: almost no one is classified positive. High specificity, very low sensitivity.

#### c = 0.50

```{r, fig.height=5, fig.width=10}
plot_cutpoint(0.50, fev$risk.score, fev$smoker, roc.demo)
```

Still a high cutpoint for this dataset. Starting to capture some true positives.

#### c = 0.30

```{r, fig.height=5, fig.width=10}
plot_cutpoint(0.30, fev$risk.score, fev$smoker, roc.demo)
```

A moderate cutpoint. Sensitivity is improving but false positives are increasing.

#### c = 0.15

```{r, fig.height=5, fig.width=10}
plot_cutpoint(0.15, fev$risk.score, fev$smoker, roc.demo)
```

A lower cutpoint. Capturing most true positives, but at the cost of more false positives.

#### c = 0.05

```{r, fig.height=5, fig.width=10}
plot_cutpoint(0.05, fev$risk.score, fev$smoker, roc.demo)
```

Very low cutpoint: nearly everyone is classified positive. High sensitivity, very low specificity.

:::


-   Key observations:

    -   A high cutpoint (e.g., $c = 0.70$) classifies very few subjects as positive — high specificity but low sensitivity

    -   A low cutpoint (e.g., $c = 0.05$) classifies almost everyone as positive — high sensitivity but low specificity

    -   The ROC curve traces out **all possible trade-offs** between sensitivity and specificity as $c$ varies

    -   A perfect classifier would have the box plots completely separated with no overlap

### Example 1: Age or height to predict smoking

-   Question: Is age or height a better predictor of smoking status?

-   Will compare the ROC curves

    -   Often summarized using the area under the curve (AUC), also called the C-statistic

    -   A null model (not predictive of outcome) will have AUC of 0.50

    -   Maximum AUC is 1.0

```{r}
# Fit two models: height alone vs age alone
m.height <- glm(smoker ~ height, data=fev, family=binomial)
m.age <- glm(smoker ~ age, data=fev, family=binomial)

# Compute ROC curves
roc.height <- roc(fev$smoker, predict(m.height, type="response"), quiet=TRUE)
roc.age <- roc(fev$smoker, predict(m.age, type="response"), quiet=TRUE)
```

```{r, fig.cap="ROC curves: Age vs Height as predictors of smoking"}
plot(roc.height, col="blue", lwd=2, main="ROC Comparison: Age vs Height")
plot(roc.age, col="red", lwd=2, add=TRUE)
legend("bottomright",
       legend=c(sprintf("Height (AUC = %.3f)", auc(roc.height)),
                sprintf("Age (AUC = %.3f)", auc(roc.age))),
       col=c("blue", "red"), lwd=2, bty="n")
```

-   Age has a visibly higher AUC than height

-   For any given false positive rate, the sensitivity is higher for age

    -   Predicted value positive would also always be higher

#### A note on formal tests comparing C-statistics

-   You may encounter formal hypothesis tests comparing two AUC values (e.g., DeLong test via `roc.test()` in `pROC`)

    -   These are commonly recommended by AI assistants and found in online tutorials

-   However, these tests have serious problems and should generally be avoided

    -   The C-statistic is an **insensitive** measure of predictive accuracy — large improvements in a model can produce tiny changes in C

    -   Comparing two C-statistics is even less informative — differences are often clinically meaningless even when statistically significant

    -   With large sample sizes, trivially small differences become "significant"

    -   The test tells you nothing about *where* the model improves or for *whom*

-   Frank Harrell has written extensively about this (see *Regression Modeling Strategies*):

    -   "It is a bad idea to use a test comparing two C-statistics to decide if one model is better than another."

    -   The C-statistic is useful as a **descriptive measure** of a single model's discriminative ability, not as a basis for model comparison

-   Better approaches for comparing models:

    -   Likelihood ratio tests (for nested models)

    -   AIC/BIC for comparing model fit

    -   Calibration assessment (does the model get the probabilities right?)

    -   Net reclassification improvement (NRI) if you must compare at specific thresholds — though this has its own issues

### Example 2: Multiple predictors of smoking

-   Will consider age, height, sex, and fev as predictors of smoking

-   Fit a multivariable logistic regression model using these covariates in a training sample

    -   Training sample will contain 60% of the observations in the current study

    -   Develop a model on the training sample, see how well it fits in the validation sample (the remaining 40% of the data)

-   Consider a rule that predicts a subject will smoke if their predicted probability is greater than 0.5

    -   Could choose other cutoffs than 0.5

    -   Other cutoffs will be represented on the ROC curves

    -   Sensitivity, specificity will vary by the cutoff chosen

```{r}
# Create training/validation split (60/40)
set.seed(42)
fev$sex.factor <- factor(fev$sex)
n <- nrow(fev)
train.idx <- sample(1:n, size=floor(0.6 * n))
fev$training <- FALSE
fev$training[train.idx] <- TRUE

# Fit multivariable model on training data
m.multi <- glm(smoker ~ age + height + sex.factor + fev, data=fev,
               subset=training, family=binomial)
summary(m.multi)
```

```{r}
# Predict on validation data
fev$pfit <- predict(m.multi, newdata=fev, type="response")
val <- fev[!fev$training, ]

# Classification table using cutoff of 0.50
val$predicted <- ifelse(val$pfit >= 0.5, 1, 0)
tab <- table(Observed=val$smoker, Predicted=val$predicted)
tab
```

```{r}
# Compute sensitivity, specificity, PVP, PVN
sens <- tab["1","1"] / sum(tab["1",])
spec <- tab["0","0"] / sum(tab["0",])
pvp <- tab["1","1"] / sum(tab[,"1"])
pvn <- tab["0","0"] / sum(tab[,"0"])

metrics <- data.frame(
  Metric = c("Sensitivity", "Specificity", "PVP", "PVN"),
  Value = c(sens, spec, pvp, pvn),
  Percent = sprintf("%.1f%%", 100 * c(sens, spec, pvp, pvn))
)
kable(metrics, digits=3,
      caption="Classification performance at cutoff = 0.50 (validation sample)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   Thresholds other than 0.50 give different values for sensitivity, specificity, PVP, PVN

    -   ROC curve displayed gives all possible thresholds

```{r, fig.cap="ROC curve for multivariable smoking prediction model (validation sample)"}
roc.multi <- roc(val$smoker, val$pfit, quiet=TRUE)
plot(roc.multi, col="blue", lwd=2,
     main=sprintf("Multivariable Model ROC (AUC = %.3f)", auc(roc.multi)))
```

### Multivariable prognostic models

-   Reference

    -   Harrell FE Jr, Lee KL, Mark DB. Multivariable prognostic models: issues in developing models, evaluating assumptions and adequacy, and measuring and reducing errors. Statistics in Medicine. 15(4): 361-87, Feb 1996

    -   A "Tutorial in Biostatistics" (frequently appears in Statistics in Medicine)

-   From abstract of paper

    -   "Multivariable regression models are powerful tools that are frequently used in clinical outcomes... however uncritical application of modeling techniques can result in models that poorly fit the dataset at hand, or, even more likely, inaccurately predict outcomes on new subjects."

    -   "... predictive accuracy should be unbiasedly validated using bootstrapping or cross-validation before using predictions in new data series."

-   Common problem: Number of predictors being considered ($p$) is larger than the number of events ($m$)

    -   Recommendation: Consider no more than $m/10$ potential predictors in a model

    -   e.g. if event rate is $10\%$ and $n=200$ then $m=20$; only $2$ potential predictors

-   Step 1: Consider your dataset

    -   Missing data (imputation?)

    -   Interactions (scientific relevance?)

    -   Transformation of predictors (scientific relevance?)

-   Step 2: Data Reduction

    -   Reduce the number of predictors to a manageable amount

    -   Utilize methods that do not consider the outcome ("unsupervised" learning)

        -   If you do not use the outcome at this stage, your statistical inference is preserved

        -   Correlations among predictors: Variable clustering, principal components analysis

        -   Scientifically meaningful summary measures

#### A note on stepwise model building

-   Old method for considering a large number of covariates that might possibly be predictive of an outcome

    -   "If stepwise model building had been proposed today, it would not have passed peer review"

-   Available in all software as an automated tool, but should not be used

-   Avoid recreating the stepwise approach when building your own models manually

-   Major caveats

    -   Overfits your dataset

    -   P-values are not true p-values; they are very anti-conservative

    -   You will often obtain different models if you use "forward" or "backward" stepwise regression

    -   Ignores confounding effects (a variable could be an important confounder, but itself not be statistically significant in the model)

    -   Without clustering certain predictors, may throw out covariates to make non-sensical model

        -   Suppose race/ethnicity is coded as White, Black, Hispanic, Other

        -   Want to model as 3 dummy variables

        -   Stepwise may throw out one of your dummy variables, keep the rest

        -   Pairwise significance would also be highly dependent on reference group chosen by analyst

-   Two flavors of stepwise model building

    -   Start with no covariates: "Forward" stepwise regression

    -   Start with all covariates: "Backward" stepwise regression

-   Stepwise procedure proceeds by adding or removing covariates from the model based on the corresponding partial $t$ or $Z$ test

    -   Must pre-specify a "P to enter" and "P to remove"

    -   To avoid infinite loops, P to enter must be less than P to remove

    -   E.g. Add a variable to the model if $p < 0.05$, but only remove a variable from the model if $p > 0.10$

    -   Repeat the process until you arrive at a final model

-   Step 3: Fit the model and ...

    -   Evaluate modeling assumptions: Linearity, additivity, distributional assumptions

    -   Use backwards stepwise selection to find a simpler model

        -   Harrell: Remove predictor if $\chi^2 < 2$ ($p > 0.16$)

    -   Will use *all data* to fit model

-   Step 4: Evaluate model

    -   Bootstrap cross-validation

        -   Sample data with replacement

        -   For each bootstrap sample, the evaluation of modeling assumptions and backwards selection will be performed (these steps use the outcome)

        -   Summarize predictive accuracy using a statistic (C-index, Brier score)

    -   Bootstrap cross-validation provides a nearly unbiased estimate of predictive accuracy (e.g. C-index, Brier score) while allowing the entire model to be used for model development

        -   C-index: Area under the ROC curve. Index near 1 indicates higher accuracy.

        -   Brier score: Average squared deviation between predicted probability of event and observed outcome. A lower score represents higher accuracy.

#### Case Study: Predicting smoking using lrm()

-   We will apply the Harrell framework to the FEV/smoking data we have been using throughout this lecture

-   **Step 1: Consider the dataset**

    -   Outcome: smoking status (binary)

    -   Available predictors: age, height, sex, FEV

    -   How many predictors can we afford?

```{r}
# Number of events (smokers) determines model complexity budget
n.events <- sum(fev$smoker)
n.total <- nrow(fev)

data.frame(
  `Total N` = n.total,
  `N smokers (events)` = n.events,
  `Event rate` = sprintf("%.1f%%", 100 * n.events / n.total),
  `Max predictors (m/10)` = floor(n.events / 10),
  check.names = FALSE
) |>
  kable(caption = "Sample size considerations for model complexity") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

-   **Step 2: Data reduction** — With only 4 candidate predictors and a reasonable number of events, we can include all predictors without formal data reduction

    -   We will allow age and height to have nonlinear effects using restricted cubic splines

    -   We will consider a sex interaction with age (biological plausibility: smoking uptake patterns may differ by sex)

-   **Step 3: Fit the model** using `lrm()` from the `rms` package

    -   `lrm()` provides richer output than `glm()`: discrimination indexes (C-statistic), likelihood ratio tests, and integration with `validate()` and `calibrate()`

```{r}
dd <- datadist(fev)
options(datadist = "dd")

f <- lrm(smoker ~ sex * rcs(age, 4) + rcs(height, 4) + fev,
         data = fev, x = TRUE, y = TRUE)
f
```

-   The model summary from `lrm()` provides several useful quantities

    -   **C-statistic**: A measure of discrimination (0.5 = coin flip, 1.0 = perfect)

    -   **Brier score**: Average squared deviation between predicted probability and observed outcome (lower is better)

    -   **$R^2$**: Nagelkerke $R^2$, a pseudo-$R^2$ measure

-   We can visualize the ANOVA-like decomposition of the model

```{r}
anova(f)
```

-   **Step 4: Evaluate the model** using bootstrap validation

    -   We fit the model using *all* the data (no train/test split)

    -   Bootstrap cross-validation provides a nearly unbiased estimate of how well the model will perform on new data

    -   The `optimism` column reveals how much we are overfitting

```{r, cache=TRUE}
v <- validate(f, B=300)
v
```

-   Key rows in the `validate()` output:

```{r}
val.guide <- data.frame(
  Index = c("Dxy", "R2", "Slope", "B"),
  Description = c(
    "Somers' Dxy rank correlation (C = 0.5 + Dxy/2)",
    "Nagelkerke R-squared",
    "Calibration slope (1.0 = perfectly calibrated)",
    "Brier score (lower is better)"
  ),
  `Original` = c("Apparent performance on full data",
                  "Apparent performance on full data",
                  "Always 1.0 on training data",
                  "Apparent Brier score"),
  `Corrected` = c("Honest estimate after removing optimism",
                   "Honest estimate after removing optimism",
                   "Degree of overfitting (< 1 means overfit)",
                   "Honest Brier score"),
  check.names = FALSE
)
kable(val.guide, caption="Guide to key validate() output") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width=FALSE)
```

-   The `optimism` column is the difference between `training` and `test` performance across bootstrap samples

    -   Large optimism indicates the model is overfitting the data

    -   The `index.corrected` = `index.orig` $-$ `optimism` gives the honest estimate

-   Calibration: Does the model get the probabilities right?

```{r, fig.cap="Bootstrap calibration plot for the smoking prediction model", cache=TRUE}
cal <- calibrate(f, B=300)
plot(cal, main="Bootstrap Calibration: Smoking Prediction Model")
```

-   The calibration plot shows predicted probability (x-axis) versus observed event rate (y-axis)

    -   A well-calibrated model follows the 45-degree line

    -   The gray region shows the confidence band from bootstrap resampling

    -   Departures from the line indicate systematic over- or under-prediction at certain risk levels

-   Summary: Discrimination tells you whether the model can *rank* patients correctly; calibration tells you whether the predicted *probabilities* are accurate. A model can have good discrimination but poor calibration (or vice versa). Both matter for clinical use.

## Summary

-   The choice of model (Lecture 12) directly affects the quality of predictions — bias from oversimplification vs. imprecision from overparameterization

-   **Prediction of summary measures** (means, probabilities):

    -   Confidence intervals quantify uncertainty about the population quantity

    -   For multivariable models, the matrix expression $\text{se}_{\text{fit}}^2 = x_h (X'X)^{-1} x_h' \cdot \text{MSE}$ makes the dependence on the design explicit

-   **Prediction of individual observations**:

    -   Prediction intervals are always wider: $\text{se}_{\text{pred}}^2 = \text{se}_{\text{fit}}^2 + \text{MSE}$

    -   The irreducible error ($\text{MSE}$) dominates in most practical settings

-   **Classification and discrimination** for binary outcomes:

    -   The ROC curve and C-statistic summarize a model's ability to rank subjects by risk

    -   Formal comparison of C-statistics across models is problematic; focus on whether discrimination is adequate for the clinical purpose

-   **ANCOVA predictions** can be conditional (at specific covariate values) or marginal (averaged over the observed distribution)

    -   Marginal means answer the population-level question; conditional means answer the individual-level question

-   **Model validation** (discrimination + calibration via bootstrapping) is essential before deploying predictions clinically
